---
title: "Module 7 - Logistic Regression"
subtitle: <center> <h1>In-Class Analysis</h1> </center>
format: 
  html:
    embed-resources: true 
---

<style type="text/css">
h1.title {
font-size: 40px;
text-align: center;
}
</style>

```{r setup, include=FALSE}
# load packages here
library(tidyverse)
library(corrplot)  # for the correlation matrix
library(bestglm)  # for variable selection
library(car)  # for the VIFs
library(pROC)  # for the ROC curve
library(ROCR)  # for the color-coded ROC curve
```

## Data and Description

Coronary heart disease (CHD) refers to a narrowing of the coronary arteries, the blood vessels that supply oxygen and blood to the heart. It is also known as coronary artery disease. It is a major cause of illness and death. CHD normally happens when cholesterol accumulates on the artery walls, creating plaques. The arteries narrow, reducing blood flow to the heart. Sometimes, a clot can obstruct the flow of blood to the heart muscle. CHD commonly causes angina pectoris (chest pain), shortness of breath, myocardial infarction, or a heart attack. It is the most common type of heart disease in the United States, where it accounts for 370,000 deaths every year.

The CHD data set contains information on 757 subjects (randomly selected) aged 39 to 59 years old and free of heart disease as determined by electrocardiogram at an initial screening. At baseline the variables in the following table were collected. Follow-up continued for 8.5 years with repeat examinations to determine if patients developed CHD. The goal is to determine risk factors (ways of healthy living) to avoid CHD.

Variable | Description
-------- | -------------
age      | Age in years
height   | Height in inches
weight   | Weight in pounds
sbp      | Systolic blood pressure in mmHg (millimeters of mercury)
dbp      | Diastolic blood pressure in mmHg (millimeters of mercury) 
chol     | Cholesterol in mg/dL (milligrams of cholesterol per deciliter of blood)
cigs     | Number of cigarettes smoked a day


Do the following:

1. Download the "CHD.csv" file from Canvas and put it in the same folder as this Quarto file.
2. Read in the data set, call it "heart", and change the class of the response (chd) to a factor. You can do this using the `mutate` function from tidyverse and `as.factor`.

```{r}
heart <- read_csv("CHD.csv") |> 
  mutate(chd = as.factor(chd))
str(heart)
```

## Explore the data. We will make just a few plots for illustration. For the age variable, create a jittered scatterplot, a boxplot, and a contingency table.

```{r, fig.align='center'}
# Jittered Scatterplot for age
ggplot(data = heart) +
  geom_point(mapping = aes(y = chd, x = cigs)) +
  geom_jitter(mapping = aes(y = chd, x = cigs),
              height = 0.1) +
  theme(aspect.ratio = 1)

# Boxplot for age
ggplot(data = heart) +
  geom_boxplot(mapping = aes(y = age, x = chd)) +
  theme(aspect.ratio = 1) +
  coord_flip()

# Cross-Tabulation (Contingency Table) for age 
table(heart$age, heart$chd)
table(heart$age) # number of people in each age level

```
## Perform some type of variable selection procedure to determine which variables to include in the model. Hint: You will need to specify the "family" argument. Hint 2: If you use cross-validation, the response variable should *not* be a factor for this function to work.

```{r, fig.align='center'}

# Does there appear to be multicollinearity based on the correlation plot?
corrplot(cor(heart[,-8]))

# For illustration, we will use best subsets (exhaustive) with the BIC metric
heart_best_subsets_bic <- bestglm(as.data.frame(heart),
                                  IC = "BIC",
                                  method = "exhaustive",
                                  TopModels = 1,
                                  family = binomial)
summary(heart_best_subsets_bic$BestModel)



library(glmnet)  # for ridge, lasso, and elastic net
set.seed(12345)  # make sure to set your seed when doing cross validation!

heart$chd_num <- ifelse(heart$chd == "yes", 1, 0)

env_x <- as.matrix(heart[, 1:7])
env_y <- heart$chd_num
# use cross validation to pick the "best" (based on MSE) lambda
env_lasso_cv <- cv.glmnet(x = env_x,
                          y = env_y,
                          family = "binomial",
                          type.measure = "mse",
                          alpha = 1)  # 1 is code for "LASSO"
coef(env_lasso_cv, s = "lambda.min")
```

## Use the `glm` function to fit a logistic regression model using the variables identified above. Note that you will need to include the following argument to the `glm` function to indicate you are performing logistic regression: `family = binomial(link = "logit")`.

```{r, fig.align='center'}
heart_logistic <- glm(chd ~ age + weight + sbp + chol + cigs,
                      data = heart,
                      family = binomial(link = "logit"))
summary(heart_logistic)

```

## Briefly, check the logistic regression model assumption: The x's vs log odds are linear (monotone in probability)

```{r, fig.align='center'}
# here is the plot for age - repeat for the other predictors
scatter.smooth(x = heart$age, y = as.numeric(heart$chd) - 1)

scatter.smooth(x = heart$weight, y = as.numeric(heart$chd) - 1)

scatter.smooth(x = heart$sbp, y = as.numeric(heart$chd) - 1)

scatter.smooth(x = heart$chol, y = as.numeric(heart$chd) - 1)

scatter.smooth(x = heart$cigs, y = as.numeric(heart$chd) - 1)



```


## Briefly, check the logistic regression model assumption: No multicollinearity, using the VIFs

```{r, fig.align='center'}
# this code uses the pseudo R-Squared
vif(heart_logistic)

# this code uses the real R-Squared (and is valid since we are looking at how
# the predictor variables relate, and we are not using the response)
heart_lm <- lm(as.numeric(chd) ~ age + weight + sbp + chol + cigs,
               data = heart)
vif(heart_lm)
max(vif(heart_lm))# < 10
mean(vif(heart_lm)) # < 5

```

## Briefly, check the logistic regression model assumption: No influential observations
```{r}
plot(heart_logistic, which = 5, cook.levels = .5) # Residuals by leverage
```


## Create confidence intervals for $\beta_k$, $\exp\{\beta_k\}$, and $100 \times (\exp\{\beta_k\} - 1)%$ for all predictors using the `confint` function. Make sure the confidence intervals for age match the ones in the notes and that you can interpret the intervals correctly.

```{r, fig.align='center'}
heart_conf_int_LOR <- confint(heart_logistic)[-1,] # omit intercept row

# A
heart_conf_int_LOR

# B
exp(heart_conf_int_LOR)

# C
100 * (exp(heart_conf_int_LOR) - 1) # Most straigtforward


```

## Calculate the predicted probability that a patient has CHD using the `predict` function for a patient where age = 50, weight = 182, sbp = 136, chol = 253, and cigs = 20.

```{r, eval=FALSE}
new_patient <- data.frame(age = 50, 
                          weight = 182, 
                          sbp = 136, 
                          chol = 253, 
                          cigs = 20)
# predicted log odds
pred_log_odds <- predict(heart_logistic, 
                         newdata = new_patient)

pred_log_odds

# predicted probability

exp(pred_log_odds)/(1 + exp(pred_log_odds))


# can do this two-step process in one step by using type = "response" to get the
# predicted probability

pred_prob <- predict(heart_logistic, 
                         newdata = new_patient,
                     type = "response")

pred_prob

```

## Calculate a 95% confidence interval for the predicted probability that a patient has CHD for a patient where age = 50, weight = 182, sbp = 136, chol = 253, and cigs = 20.

```{r, fig.align='center'}
new_patient <- data.frame(age = 50, 
                          weight = 182,
                          sbp = 136, 
                          chol = 253,
                          cigs = 20)

# get the log odds ratio with the standard error

log_odds <- predict(heart_logistic,
                    newdata = new_patient,
                    se.fit = TRUE)

# compute the margin of error

moe <- qnorm(p = .975, lower.tail = TRUE) * log_odds$se.fit
moe

# compute the 95% confidence interval (and point estimate) for the log odds

pred_interval <- log_odds$fit + c(-1, 0, 1) * moe

# compute the 95% confidence interval (and point estimate) for the predicted
# probability

exp(pred_interval) / (1 + exp(pred_interval))
```



## Check the logistic regression model performance

### Compute the likelihood ratio test statistic (aka the deviance, aka the model chi-squared test) and its corresponding p-value. Hint: use the deviances reported in the logistic regression model output for the test statistic, and use the `pchisq` function for the p-value.

```{r, fig.align='center'}
# Likelihood ratio test statistic
like_ratio <- heart_logistic$null.deviance - heart_logistic$deviance

# Likelihood ratio p-value
pchisq(q = like_ratio,
       df = length(coef(heart_logistic)) - 1,
       lower.tail = FALSE)


```

### Compute the Pseudo R-Squared. Hint: use the deviances reported in the logistic regression model output.

```{r, fig.align='center'}
1 - heart_logistic$deviance/heart_logistic$null.deviance

```

### Find the best cutoff value for classification by minimizing the percent misclassified (c in the notes).

```{r, fig.align='center'}
# get the predicted probabilities for all 757 patients:
heart_preds <- predict(heart_logistic,
                       type = "response")

# Look at the first few predictions:

head(heart_preds)
# create a sequence from 0 to 1 to represent all possible cut-off values (c) 
# that we could choose:

possible_cutoffs <- seq(0, 1, by = .01)


# transform heart$chd from a factor with levels "yes" and "no" to a numeric of 
# 1's and 0's:

heart_binary <- ifelse(heart$chd == "yes", 1, 0)
heart_binary

# create an empty vector where we will store the percent misclassified for each
# possible cut-off value we created:

percent_missclass <- rep(NA, length(possible_cutoffs))


# for each possible cut-off value, (1) grab the cut-off value, (2) for all 757
# patients, store a 1 in "classify" if their predicted probability is larger 
# than the cut-off value, and (3) compute the average percent misclassified 
# across the 757 patients when using that cut-off by averaging the number of 
# times "classify" (0 or 1 based on how that cut-off classified a person) is 
# not the same as heart_binary (the truth):

for(i in 1:length(possible_cutoffs)){
  classify <- ifelse(heart_preds > possible_cutoffs[i], 1, 0)
  percent_missclass[i] <- mean(classify != heart_binary)
}


# percent_misclass holds the average misclassification rates for each cut-off
percent_missclass

# put this information in a dataframe so we can plot it with ggplot:
missclass_df <- as.data.frame(cbind(percent_missclass, possible_cutoffs))

# plot the misclassification rate against the cut-off value:
ggplot(data = missclass_df) +
  geom_line(aes(x = possible_cutoffs, y = percent_missclass))

# choose the "best" cut-off that minimizes the percent misclassified:

cutoff_best <- possible_cutoffs[which.min(percent_missclass)]

percent_missclass == min(percent_missclass) # .53 and .54 also minimize

```

### Create a confusion matrix based on the best cutoff (note that the order is changed in the table created here vs the table in the notes)

```{r,fig.align='center'}
# for each of the 757 patients, compare their predicted probability with the
# chosen cut-off value. If their predicted probability is larger than the 
# cut-off store "TRUE", otherwise, store "FALSE" in preds

preds <- heart_preds > cutoff_best

# create a confusion matrix with the truth and the predicted classification: 

conf_mat <- table("truth" = heart$chd, "predicted" = preds)
conf_mat

# note that we can also add column and row sums, which is useful for 
# calculations:

addmargins(conf_mat)


```

### Plot the ROC curve and obtain the AUC

```{r,fig.align='center'}
# METHOD 1: using the pROC package
my_roc <- roc(heart$chd, heart_preds)

ggplot() +
  geom_path(mapping = aes(x = 1 - my_roc$specificities, 
                          y = my_roc$sensitivities), 
            linewidth = 2) +
  geom_abline(slope = 1, intercept = 0) +
  theme_bw() + 
  xlab("1 - Specificity (False Positive Rate)") +
  ylab("Sensitivity (True Positive Rate)") +
  theme(aspect.ratio = 1)

auc(my_roc)  # AUC

# METHOD 2: using the ROCR package (ROC curve colored based on cutoff value)
pred <- prediction(heart_preds, heart$chd)
perf <- performance(pred, "tpr", "fpr")  # tpr = true positive rate, 
# fpr = false positive rate
plot(perf, 
     colorize = TRUE, 
     lwd = 6,
     xlab = "1 - Specificity (False Positive Rate)",
     ylab = "Sensitivity (True Positive Rate)")
abline(a = 0, b = 1)

# AUC
auc <- performance(pred, measure = "auc")
auc@y.values[[1]]

str(auc)
```


## Summary and Conclusions

Logistic regression is used when you have a binary response variable. Many of the things we learned for linear regression apply to logistic regression, but there are several differences. For instance, we have few model assumptions, we measure performance based on deviances (instead of sums of squares), the coefficients are in terms of log odds ratios, and we can perform classification. 

