---
title: "Module 3 - Simple Linear Regression Model Inference"
subtitle: <center> <h1>In-Class Analysis</h1> </center>
format: 
  html:
    embed-resources: true 
---

```{=html}
<style type="text/css">
h1.title {
  font-size: 40px;
  text-align: center;
}
</style>
```

```{r setup, include=FALSE}
# load packages here
library(tidyverse)
```

## Data

This in-class analysis uses the Windmill data set that we have used previously in Module 1. The Windmill data set contains measurements of wind speed (in meters per second m/s) at a **candidate site (CSpd) (column 1)** and at an accompanying **reference site (RSpd) (column 2)** for 1,116 areas. Download the Windmill.txt file from Canvas, and put it in the same folder as this quarto file.

Do the following:

1.  Read in the data set, assign it to a variable `wind`, and take a look at the top 15 rows.
2.  Create a scatterplot of the data, and overlay the fitted linear regression line.
3.  Apply linear regression to the data with `CSpd` as response and `RSpd` as predictor.
4.  Save the model fitted values $\hat{y}$ and the residuals.
5.  Examine the summary of the model fit using the `summary` function. (For help on this function, type `?summary.lm` in your R console.)

```{r}
# When performing step 3, use the format lm(y ~ x, data = wind), where
# y and x are the variable names for the response and predictor
windmill_data <- read_table("Windmill.txt")

plot <- ggplot(data = windmill_data, mapping = aes(x = RSpd, y = CSpd)) +
  geom_point(color = "blue", size = 1) +
  labs(title = "The Relationship Between a Car's Speed and Stopping Distance",
       x = "RSpd",
       y = "CSpd") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_minimal()

print(plot)
```

```{r}
print(summary(linear_model))
```

## Statistical Inference for the Slope

#### Using the formula from class, compute $\mathrm{\widehat{Std.Err.}}(\hat{\beta}_1)$, the standard error of the OLS estimator for the slope. Check your answer by comparing with the standard error reported in the summary above.

```{r}
residuals <- linear_model$residuals

std_error_regression_line <- std(residuals) * sqrt((1/length(residuals)) * ((xstar - mean(windmill_data$RSpd))^2 / sum((windmill_data$RSpd - mean(windmill_data$RSpd))^2)
```

\< your response here \>

#### Use the `confint` R function to create a 99% confidence interval for the slope. How would you interpret this interval? What would you expect to change about this interval if the level was changed to 90%, and why?

```{r}
# Uncomment the code below to generate the CI. You must replace my_lm with
# the name of the variable to which you assigned the linear model fit
# If you remove 'par = 2', you will get CIs for the intercept and slope instead
# of just the slope. You can also use par = "RSpd".


#CI_slope_99 <- confint(my_lm, par = 2, level = 0.99)
#CI_slope_99
```

\< your response here \>

#### Use the fitted model output and the R function `pt()` to conduct a two-sided hypothesis test on the slope, that the slope is not equal to 0. The test statistic and p-value you obtain should match those in the output from `summary`. \*Note that the `summary` output results from a two-sided test. Use `?pt` if you are not sure what this function does.

```{r}
# < your code here >
```

\< your response here \>

## Statistical Inference for the Mean Response

#### Using the `predict` function, calculate a 90% confidence interval for the average wind speed at the candidate site when the measured wind speed at the reference site is 8 m/s.

```{r}
# Uncomment the code below to create the CI, again replacing my_lm with the 
# appropriate variable name
#CI_90_x8 <- predict(my_lm, newdata = data.frame(RSpd = 8), interval = 'confidence')
#CI_90_x8
```

#### Create a confidence band for all average wind speeds at hypothetical candidate sites corresponding to all reference site wind speeds within the range of the data.

```{r}
# If wind.base.plot is your base scatterplot of wind speeds (not including the 
# fitted regression line), the following code will produce the desired plot

# wind.base.plot + geom_smooth(method = 'lm', se = TRUE)
```

## Statistical Inference for the Individual Observations

#### Using the `predict` function, calculate an 90% prediction interval for the wind speed at a future candidate site when the wind speed at the reference cite is 8 m/s. Explain why, even though the level is the same as the level used above for the confidence interval of the mean response, the prediction interval is much wider.

```{r}
# < your code here >
```

#### Using the `predict` function with a data frame of many $x$ values (use the `seq` R function starting at the minimum of `RSpd` and going to the maximum of `RSpd`), create a prediction band for candidate site wind speeds across all values of reference site wind speeds. Overlay this prediction band on your previous scatterplot using the `geom_line` function twice.

```{r}
# The code below will generate the desired band
#
# RSpd_grid <- seq(min(wind$RSpd), max(wind$RSpd), length.out = 100)
# wind_PI <- predict(my_lm, level = 0.9, 
#         newdata = data.frame(RSpd = RSpd_grid), 
#         interval = "prediction") |> 
#   as_tibble() |> 
#   mutate(RSpd = RSpd_grid) |> 
#   rename("CSpd" = fit)
# 
# wind.base.plot + 
#   geom_ribbon(data = wind_PI, 
#               aes(ymin = lwr, ymax = upr),
#               alpha = 0.25, # This just changes the opacity of the band
#               outline.type = 'both') +
#     geom_smooth(method = 'lm', se = FALSE) 
```

## ANOVA for Regression

#### Using only the lm model summary output above (i.e., without using the `aov()` or `anova()` functions), complete the ANOVA table below for your fitted linear regression model (places with a `?` should be replaced by your answer). **Once you are finished, use R to construct the ANOVA table and verify your calculations**.

| Source   | Degrees of Freedom | Sum of Squares | Mean Square |
|----------|:------------------:|:--------------:|:-----------:|
| RSpd     |         ?          |     ??????     |   9015.6    |
| Residual |        ????        |    6775.75     |     ???     |
| Total    |        1115        |    15791.34    |             |

```{r}
# < your code here >
```

#### What value in the ANOVA table gives the variance estimate $\hat{\sigma}^2$ (or $s_e^2$).

\< your response here \>

#### Calculate R-Squared (Coefficient of Determination) using the values from the ANOVA table above. The value you get should match the value from the `summary` function you used above.

```{r}
# <your code here>
```

#### Recall that the total sum of squares (SST, also called $S_{YY}$) is $\sum_{i = 1}^n (Y_i - \overline{Y})^2.$ Using the fitted model, verify that

$$
SSE = SST(1 - R^2).
$$ Since $SST$ is the total variability in the $Y_i$s, and $SSE$ is the variability left in the residuals *after* regression is performed, $R^2$ can be interpreted as the **proportion of variability in the response explained by a linear relationship with the response**.

```{r}
# <your code here>
```
