---
title: "Module 3 - Simple Linear Regression Model Inference"
subtitle: <center> <h1>In-Class Analysis</h1> </center>
format: 
  html:
    embed-resources: true 
---

```{=html}
<style type="text/css">
  h1.title {
    font-size: 40px;
    text-align: center;
  }
</style>
```

```{r setup, include=FALSE}
# load packages here
library(tidyverse)
library(car)
```

## Data

This in-class analysis uses the Windmill data set that we have used previously in Module 1. The Windmill data set contains measurements of wind speed (in meters per second m/s) at a **candidate site (CSpd) (column 1)** and at an accompanying **reference site (RSpd) (column 2)** for 1,116 areas. Download the Windmill.txt file from Canvas, and put it in the same folder as this quarto file.

Do the following:

1.  Read in the data set, assign it to a variable `wind`, and take a look at the top 15 rows.
2.  Create a scatterplot of the data, and overlay the fitted linear regression line.
3.  Apply linear regression to the data with `CSpd` as response and `RSpd` as predictor.
4.  Save the model fitted values $\hat{y}$ and the residuals.
5.  Examine the summary of the model fit using the `summary` function. (For help on this function, type `?summary.lm` in your R console.)

```{r}
# When performing step 3, use the format lm(y ~ x, data = wind), where
# y and x are the variable names for the response and predictor

# 1. 
wind <- read_table("Windmill.txt")

# 2.
ggplot(data = wind, aes(RSpd, CSpd)) +
  geom_point() +
  geom_smooth(se = F)

# 3. 
wind_lm <- lm(CSpd ~ RSpd, data = wind)

# 4.
resids <- wind_lm$residuals
y_hat <- wind_lm$fitted.values

# 5.
summary(wind_lm)

```

## Statistical Inference for the Slope

#### Using the formula from class, compute $\mathrm{\widehat{Std.Err.}}(\hat{\beta}_1)$, the standard error of the OLS estimator for the slope. Check your answer by comparing with the standard error reported in the summary above.

```{r}
S_XX <- sum((wind$RSpd - mean(wind$RSpd))^2)
s_e <- sqrt(sum(resids^2/(nrow(wind) - 2)))
# OR
s_e <- summary(wind_lm)$sigma

(b1_hat_std_err <- s_e / sqrt(S_XX))

```

The value $\mathrm{\widehat{SE}}(\hat{\beta}_1) =$ `r round(b1_hat_std_err, digits = 4)` matches the value from the summary output above.

#### Use the `confint` R function to create a 99% confidence interval for the slope. How would you interpret this interval? What would you expect to change about this interval if the level was changed to 90%, and why?

```{r}
# Uncomment the code below to generate the CI. You must replace my_lm with
# the name of the variable to which you assigned the linear model fit
# If you remove 'par = 2', you will get CIs for the intercept and slope instead
# of just the slope. You can also use par = "RSpd".


CI_slope_99 <- confint(wind_lm, par = 2, level = 0.99)
CI_slope_99
```

We are 99% confident that the change in average wind speed at the candidate site will be between 0.71 and 0.81 m/s when the wind speed at the reference site increases by 1 m/s. The interval would shrink somewhat if the level was changed to 90%. Intuitively, this happens because we are imposing a lower level of confidence. Mathematically, the t-percentile in the CI formula would decrease, which would then decrease the length of the interval.

#### Use the fitted model output and the R function `pt()` to conduct a two-sided hypothesis test on the slope, that the slope is not equal to 0. The test statistic and p-value you obtain should match those in the output from `summary`. \*Note that the `summary` output results from a two-sided test. Use `?pt` if you are not sure what this function does.

```{r}
tstat <- wind_lm$coefficients[2] / b1_hat_std_err
pval <- 2 * pt(abs(tstat), 
               df = nrow(wind) - 2,
               lower.tail = FALSE)
pval
```

The p-value is extremely small, so we reject the null hypothesis and conclude that the average wind speed at a candidate site has a statistically significant linear relationship with wind speeds at the reference sites.

## Statistical Inference for the Mean Response

#### Using the `predict` function, calculate a 90% confidence interval for the average wind speed at the candidate site when the measured wind speed at the reference site is 8 m/s.

```{r}
# Uncomment the code below to create the CI, again replacing my_lm with the 
# appropriate variable name
CI_90_x8 <- predict(wind_lm, 
                    newdata = data.frame(RSpd = 8), 
                    interval = 'confidence',
                    level = .9)
CI_90_x8
```

#### Create a confidence band for all average wind speeds at hypothetical candidate sites corresponding to all reference site wind speeds within the range of the data.

```{r}
# If wind.base.plot is your base scatterplot of wind speeds (not including the 
# fitted regression line), the following code will produce the desired plot

wind.base.plot <- ggplot(data = wind, aes(RSpd, CSpd)) +
  geom_point()

wind.base.plot + 
  geom_smooth(method = 'lm', 
              se = TRUE,
              level = .9) # for a 90%  CI

```

## Statistical Inference for the Individual Observations

#### Using the `predict` function, calculate an 90% prediction interval for the wind speed at a future candidate site when the wind speed at the reference cite is 8 m/s. Explain why, even though the level is the same as the level used above for the confidence interval of the mean response, the prediction interval is much wider.

```{r}
PI_90_x8 <- predict(wind_lm, 
                    newdata = data.frame(RSpd = 8), 
                    interval = 'predict',
                    level = .9)
PI_90_x8
```

The interval is wider because it is for a single measurement of wind speed at the candidate site, as opposed to the average wind speed at that candidate site.

#### Using the `predict` function with a data frame of many $x$ values (use the `seq` R function starting at the minimum of `RSpd` and going to the maximum of `RSpd`), create a prediction band for candidate site wind speeds across all values of reference site wind speeds. Overlay this prediction band on your previous scatterplot using the `geom_line` function twice.

```{r}

RSpd_grid <- seq(min(wind$RSpd), max(wind$RSpd), length.out = 100)
wind_PI <- predict(wind_lm, 
                   level = 0.9,
                   newdata = data.frame(RSpd = RSpd_grid),
                   interval = "prediction") |>
  as_tibble() |>
  mutate(RSpd = RSpd_grid) |>
  rename("CSpd" = fit)

wind.base.plot +
    geom_smooth(method = 'lm', se = FALSE) + 
  geom_line(data = wind_PI, aes(y = lwr), color = "red") +
  geom_line(data = wind_PI, aes(y = upr), color = "red")

# The code below will generate the desired band as well

wind.base.plot +
  geom_ribbon(data = wind_PI,
              aes(ymin = lwr, ymax = upr),
              alpha = 0.25, # This just changes the opacity of the band
              outline.type = 'both') +
    geom_smooth(method = 'lm', se = FALSE)
```

## ANOVA for Regression

#### Using only the lm model summary output above (i.e., without using the `aov()` or `anova()` functions), complete the ANOVA table below for your fitted linear regression model (places with a `?` should be replaced by your answer). **Once you are finished, use R to construct the ANOVA table and verify your calculations**.

| Source   | Degrees of Freedom | Sum of Squares | Mean Square |
|----------|:------------------:|:--------------:|:-----------:|
| RSpd     |         1          |     9015.6     |   9015.6    |
| Residual |        1114        |    6775.75     |    6.08     |
| Total    |        1115        |    15791.34    |             |

```{r}
anova(wind_lm)
```

#### What value in the ANOVA table gives the variance estimate $\hat{\sigma}^2$ (or $s_e^2$)?

The MSE or Mean Square of the Residuals. It is 6.08.

#### Calculate R-Squared (Coefficient of Determination) using the values from the ANOVA table above. The value you get should match the value from the `summary` function you used above.

```{r}
R_sq <- 1 - (6775.75/15791.34) 
R_sq
summary(wind_lm)$r.squared

# Difference is due to rounding error.
```

#### Recall that the total sum of squares (SST, also called $S_{YY}$) is $\sum_{i = 1}^n (Y_i - \overline{Y})^2.$ Using the fitted model, verify that

$$
  SSE = SST(1 - R^2).
$$ Since $SST$ is the total variability in the $Y_i$s, and $SSE$ is the variability left in the residuals *after* regression is performed, $R^2$ can be interpreted as the **proportion of variability in the response explained by a linear relationship with the response**.

```{r}
SSE <- sum(resids^2)
SSE

15791.34 * (1 - R_sq)

# Difference is due to rounding error.
```
