---
title: "Binary Outcomes I: 2×2 Tables and Tests"
subtitle: "Stat 437"
author: "Philip White"
format:
  revealjs:
    slide-number: true
    incremental: true
    theme: serif
    css: slides.css
    scrollable: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Lecture Goals

In this lecture we will:

-   Discuss the 2×2 table\
-   Derive expected counts under independence\
-   Derive and compute Pearson’s chi-square test\
-   Discuss Yates continuity correction\
-   Derive the Wald test for $\log(OR)$\
-   Derive Fisher’s exact test probability\
-   Connect these tests conceptually to logistic regression

These notes are designed so you can compute everything by hand.

------------------------------------------------------------------------

## The 2×2 Table

Observed counts:

|              | $Y=1$ | $Y=0$ | Row total |
|-------------:|:-----:|:-----:|:---------:|
|        $X=1$ |  $a$  |  $b$  |   $a+b$   |
|        $X=0$ |  $c$  |  $d$  |   $c+d$   |
| Column total | $a+c$ | $b+d$ |    $n$    |

Total sample size:

$$n = a+b+c+d$$

Risks:

$$\widehat p_1 = \frac{a}{a+b}, \quad \widehat p_0 = \frac{c}{c+d}$$

Odds ratio:

$$\widehat{OR} = \frac{a d}{b c}$$

------------------------------------------------------------------------

## The RHC Dataset: Hello Darkness, My Old Friend

The RHC dataset comes from an observational study of critically ill
patients.

-   Exposure ($X$): Right Heart Catheterization (RHC)

-   $X = 1$ → patient received RHC

-   $X = 0$ → patient did not receive RHC

-   Outcome ($Y$): In-hospital survival

-   $Y = 1$ → died

-   $Y = 0$ → survived

------------------------------------------------------------------------

## Why is this dataset interesting?

-   RHC is an invasive monitoring procedure used in ICU settings.

-   Key question:

> Does RHC increase mortality?\
> Or are sicker patients more likely to receive RHC?

-   This is an **observational dataset**, not randomized.

-   That means:

-   Treatment assignment was not random

-   Patients receiving RHC may differ systematically from others

-   Naive comparisons may reflect confounding

------------------------------------------------------------------------

## For Today

In this lecture, we will:

-   Ignore confounding temporarily
-   Treat this as a simple 2×2 problem
-   Test for association between RHC and death

Later lectures will revisit this dataset to adjust for confounding.

------------------------------------------------------------------------

## Worked Example: RHC Dataset

|              | $Y=1$ | $Y=0$ | Row total |
|-------------:|:-----:|:-----:|:---------:|
|        $X=1$ |  $a$  |  $b$  |   $a+b$   |
|        $X=0$ |  $c$  |  $d$  |   $c+d$   |
| Column total | $a+c$ | $b+d$ |    $n$    |

```{r}
library(ATbounds)   # RHC dataset
data(RHC)

tab_raw <- table(RHC$RHC, RHC$survival)

tab <- tab_raw[c("1","0"), c("1","0")]
rownames(tab) = c("X = 1", "X = 0")
colnames(tab) = c("Y = 1", "Y = 0")

tab

a <- tab[1,1]
b <- tab[1,2]
c <- tab[2,1]
d <- tab[2,2]
n <- sum(tab)
```

------------------------------------------------------------------------

## Null Hypothesis Statements

We test:

$$H_0: P(Y=1 \mid X=1) = P(Y=1 \mid X=0)$$

$$H_0: \textrm{Is the probability of survival the same with and without RHC?}$$

Equivalent formulations:

$$p_1 = p_0$$

$$OR = 1$$

$$\log(OR) = 0$$

$$P(X,Y) = P(X)P(Y)$$

All procedures in this lecture test this same null.

------------------------------------------------------------------------

## The Stat 121 Approaches

-   Two-sample Proportion Test
-   $\chi$-squared Test for Independence

------------------------------------------------------------------------

## The Two-Sample Proportion Test

|              | $Y=1$ | $Y=0$ | Row total |
|-------------:|:-----:|:-----:|:---------:|
|        $X=1$ |  $a$  |  $b$  |   $a+b$   |
|        $X=0$ |  $c$  |  $d$  |   $c+d$   |
| Column total | $a+c$ | $b+d$ |    $n$    |

Suppose we observe two independent samples:

-   Group 1: $n_1 = a+b$ observations, with $a$ successes\

-   Group 0: $n_0 = c+d$ observations, with $c$ successes

-   Sample proportions:
    $$\widehat{p}_1= \frac{a}{a + b}= \frac{a}{n_1}, \qquad 
    \widehat{p}_0 = \frac{c}{c + d}= \frac{c}{n_0}$$

------------------------------------------------------------------------

### Null Hypothesis

$$
H_0: p_1 = p_0
$$

Under $H_0$, both groups share a common probability:

$$
p = p_1 = p_0
$$

We estimate this common probability using the pooled estimator:

$$
\widehat{p} = \frac{a + c}{a + b + c + d} = \frac{a + c}{n_0 + n_1} = \frac{a + c}{n}
$$

For later,

\begin{align*}
n\widehat{p} &= a + c \\
n\left(1 - \widehat{p}\right) &= b + d \\
\end{align*}

------------------------------------------------------------------------

### Variance Under the Null

Under $H_0$, the variance of the difference in sample proportions is:

$$
\mathrm{Var}(\widehat p_1 - \widehat p_0)
=
\widehat p (1-\widehat p)
\left(
\frac{1}{n_1} + \frac{1}{n_0}
\right)
$$

-   This uses the pooled estimate because we are assuming $p_1 = p_0$
    under the null.

-   So, the standard error is

$$
\mathbb{SE}(\widehat p_1 - \widehat p_0)
=
\left(\widehat p (1-\widehat p)
\left(
\frac{1}{n_1} + \frac{1}{n_0}
\right)\right)^{1/2}
$$

------------------------------------------------------------------------

### Test Statistic

$$
Z =
\frac{\widehat p_1 - \widehat p_0}
{\left(
\widehat p(1-\widehat p)
\left(
\frac{1}{n_1} + \frac{1}{n_0}
\right)
\right)^{1/2}}
$$

Under $H_0$ and large samples:

$$
Z \sim N(0,1)
$$

Two-sided p-value:

$$
p = 2(1-\Phi(|Z|))
$$

------------------------------------------------------------------------

### Confidence Interval for $p_1 - p_0$

For confidence intervals (not testing), it's more common to use the
**unpooled variance**:

$$
SE(\widehat p_1 - \widehat p_0)
=
\left(
\frac{\widehat p_1(1-\widehat p_1)}{n_1}
+
\frac{\widehat p_0(1-\widehat p_0)}{n_0}
\right)^{1/2}
$$

95% CI:

$$
(\widehat p_1 - \widehat p_0)
\pm
1.96 \cdot SE(\widehat p_1 - \widehat p_0)
$$

Note the important distinction:

-   **Test** uses pooled variance\
-   **Confidence interval** uses unpooled variance

------------------------------------------------------------------------

## Expected Counts Under Independence

Under independence:

$$P(X=i, Y=j) = P(X=i)P(Y=j)$$

Expected counts:

$$E_{11} = \frac{(a+b)(a+c)}{n}$$

$$E_{12} = \frac{(a+b)(b+d)}{n}$$

$$E_{21} = \frac{(c+d)(a+c)}{n}$$

$$E_{22} = \frac{(c+d)(b+d)}{n}$$

These preserve row and column totals.

------------------------------------------------------------------------

## Pearson Chi-Square Statistic

$$\chi^2 = \sum \frac{(O - E)^2}{E}$$

Explicitly:

\begin{align*}
\chi^2 &=
\frac{\left(a-E_{11}\right)^2}{E_{11}} +
\frac{\left(b-E_{12}\right)^2}{E_{12}} +
\frac{\left(c-E_{21}\right)^2}{E_{21}} +
\frac{\left(d-E_{22}\right)^2}{E_{22}} \\
& = \frac{n(ad - bc)^2}{(a+b)(c+d)(a+c)(b+d)} \tag{1}.
\end{align*}

That second simplification takes "a bit" of algebra.

Under $H_0$:

$$\chi^2 \sim \chi^2_1$$

------------------------------------------------------------------------

## Using Pearson Chi-Square Statistic

Interpretation:

-   If observed counts are close to expected counts → small $\chi^2$
-   If observed counts differ substantially from expected → large
    $\chi^2$
-   Reject $H_0$ when $\chi^2$ exceeds critical value

Rule of thumb:

-   Expected counts should be $\geq 5$ in each cell for reliable
    approximation.

------------------------------------------------------------------------

## A Few Preliminaries

-   Note,
    $\widehat p_1 - \widehat p_0= \frac{a}{a+b} - \frac{c}{c+d}= \frac{a(c+d) - c(a+b)}{(a+b)(c+d)}= \frac{ad - bc}{(a+b)(c+d)}.$

-   So, $ad - bc = (\widehat p_1 - \widehat p_0)(a+b)(c+d).$

-   Also,
    $\frac{1}{a+b} + \frac{1}{c+d} = \frac{(a+b) + (c+d)}{(a+b)(c+d)}= \frac{n}{(a+b)(c+d)}.$

-   This implies that
    $\frac{(a+b)(c+d)}{n} = \frac{1}{\dfrac{1}{a+b} + \dfrac{1}{c+d}} = \frac{1}{\dfrac{1}{n_1} + \dfrac{1}{n_0}}.$

------------------------------------------------------------------------

## Equivalence to the Two-Proportion Z-Test

\begin{align*}
\chi^2 &= \frac{n(ad - bc)^2}{(a+b)(c+d)(a+c)(b+d)}  &\\
&= \frac{n\left((\widehat p_1 - \widehat p_0)(a+b)(c+d)\right)^2}{(a+b)(c+d)(a+c)(b+d)}  & \textrm{from preliminaries}\\
&= \frac{n(\widehat p_1 - \widehat p_0)^2 (a+b)(c+d)}{(a+c)(b+d)}  & \textrm{by cancellation}\\
&= \frac{n(\widehat p_1 - \widehat p_0)^2 (a+b)(c+d)}{n\widehat{p} n(1 - \widehat{p})}  & \textrm{from Null Hypothesis slide} \\
&= \frac{(\widehat p_1 - \widehat p_0)^2 (a+b)(c+d)}{n\widehat{p}(1 - \widehat{p})}  & \textrm{from Null Hypothesis slide} \\
&= \frac{(\widehat p_1 - \widehat p_0)^2}{\widehat{p}(1 - \widehat{p}) \left(\dfrac{1}{n_1} + \dfrac{1}{n_0}\right) }  & \textrm{from preliminaries} \\
&= Z^2 & \textrm{from Test Statistic Slide}\\
\end{align*}

------------------------------------------------------------------------

## Who Cares? Why This Equivalence Matters

We just showed:

$$\chi^2 = Z^2$$

in the 2×2 case.

What does this mean?

------------------------------------------------------------------------

## Proportion Test in `R`

-   

```{r}
tab

prop.test(tab,correct = FALSE)
```

------------------------------------------------------------------------

## Yates Continuity Correction

For 2×2 tables, Pearson’s statistic may overstate evidence when counts
are small.

Yates correction adjusts:

$$\chi^2_Y = \sum \frac{(|O - E| - 0.5)^2}{E}$$

This reduces type I error in small samples.

In R:

-   `prop.test()` uses Yates correction by default.
-   Set `correct = FALSE` to match Pearson exactly.

------------------------------------------------------------------------

## Proportion Test in `R` w/ Yates Correction

```{r}
tab

prop.test(tab)
```

------------------------------------------------------------------------

## Wald Test for $\log(OR)$ — Why Work on the Log Scale?

We work on the log scale because:

-   The sampling distribution of $\log(\widehat{OR})$ is approximately
    normal.
-   The log transformation makes the distribution more symmetric.
-   The delta method, which you probably aren't familiar with, gives a
    simple variance formula.

Estimator:

$$
\widehat{\log OR} = \log\left(\frac{a d}{b c}\right)
$$

------------------------------------------------------------------------

### Standard Error (Delta Method)

Approximate standard error:

$$
SE(\widehat{\log OR}) =
\left(
\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}
\right)^{1/2}
$$

Where does this come from?

-   The cell counts are approximately binomial.
-   We apply the delta method to the function $\log(ad/bc)$.
-   Delta Method: Compute the variance of the Taylor series expansion.
-   Variances of independent counts add.

Important:

> -   This approximation works well when all cell counts are reasonably
>     large.

-   If any of $a,b,c,d$ are small, the Wald approximation may be
    unstable.

------------------------------------------------------------------------

## Wald Test for $\log(OR)$ — Inference and Interpretation

Wald statistic:

$$
Z = \frac{\widehat{\log OR}}{SE(\widehat{\log OR})}
$$

Under $H_0$ (large samples):

$$
Z \sim N(0,1)
$$

Two-sided p-value:

$$
p = 2(1-\Phi(|Z|))
$$

------------------------------------------------------------------------

### Confidence Interval

95% CI on log odds scale:

$$
\widehat{\log OR} \pm 1.96 \cdot SE(\widehat{\log OR})
$$

Exponentiate to obtain CI for OR:

$$
\exp\left(
\widehat{\log OR} \pm 1.96 \cdot SE(\widehat{\log OR})
\right)
$$

### Interpretation

-   If the CI for OR excludes 1 → statistically sig nificant.
-   If OR \> 1 → exposure associated with higher odds of outcome.
-   If OR \< 1 → exposure associated with lower odds.

------------------------------------------------------------------------

## Logistic Regression Connection

Model:

\begin{align*}
\log\left(
\frac{P(Y=1 | X = x)}{1-P(Y=1| X = x)}
\right) &= \beta_0 + \beta_1 x \\
\end{align*}

-   For binary $X$: $$\beta_1 = \log(OR)$$

-   Testing: $$H_0: \beta_1 = 0$$

-   We will flesh this out in future lectures.

-   I generally think this is cleaner than a Wald Test alone

-   The Wald test can be used within logistic regression.

------------------------------------------------------------------------

## The Score (Rao) Test in Logistic Regression

In logistic regression, we test:

$$
H_0: \beta_1 = 0
$$

There are three common large-sample tests:

-   **Wald test** — uses $\widehat{\beta}_1$ and its estimated standard
    error\
-   **Likelihood ratio test (LRT)** — compares full vs reduced models\
-   **Score (Rao) test** — evaluates the slope of the log-likelihood at
    the null

------------------------------------------------------------------------

### What is the Score Test?

The score test asks:

> If the null model ($\beta_1 = 0$) were true, how steeply does the
> log-likelihood increase if we move away from 0?

It only requires fitting the **null model**.

The statistic has the general form (Details beyond scope of class):

$$
S = \frac{U(0)^2}{I(0)}
$$

where:

-   $U(0)$ = score (first derivative of log-likelihood) at $\beta_1 = 0$
-   $I(0)$ = Fisher information at $\beta_1 = 0$

------------------------------------------------------------------------

## Another Equivalence

In the 2×2 case:

-   The logistic regression score test
-   The Pearson chi-square test

are algebraically equivalent.

------------------------------------------------------------------------

## Fisher’s Exact Test — Why Do We Need It?

Pearson’s chi-square test or the two-proportion $z$-test relies on a
large-sample approximation:

$$
\chi^2 \sim \chi^2_1 \quad \text{(approximately)}
$$

This approximation works well when:

-   Expected counts in each cell are reasonably large (rule of thumb: ≥
    5)  
-   Sample size is moderate to large

When counts are small:

-   The chi-square approximation may inflate type I error.
-   The Wald confidence interval for the odds ratio can be unstable

Fisher’s exact test avoids asymptotics entirely.

It computes the **exact sampling distribution** under the null.

------------------------------------------------------------------------

## Fisher’s Exact Test — The Hypergeometric Distribution

|              | $Y=1$ | $Y=0$ | Row total |
|-------------:|:-----:|:-----:|:---------:|
|        $X=1$ |  $a$  |  $b$  |   $a+b$   |
|        $X=0$ |  $c$  |  $d$  |   $c+d$   |
| Column total | $a+c$ | $b+d$ |    $n$    |

-   Fisher’s test conditions on row and column totals

-   Given fixed margins, the only random quantity is the cell count $a$.

-   Under the null hypothesis of independence ($P(X,Y) = P(X)P(Y)$): $$
    P(a) =
    \frac{\binom{a+b}{a}\binom{c+d}{c}}
    {\binom{n}{a+c}}
    $$

-   This is the hypergeometric probability:

-   From $n$ total observations,

-   $a+c$ are "successes" (Y = 1),

-   We draw $a+b$ observations (X = 1),

-   The probability that exactly $a$ of them are successes is given by
    the expression above.

-   No approximations are used.

------------------------------------------------------------------------

## Fisher’s Exact Test — Computing the P-value

-   The exact p-value is computed by:

1.  Enumerating all possible 2×2 tables with the same margins.
2.  Computing the hypergeometric probability for each possible table.
3.  Summing the probabilities of tables that are:

-   As extreme or more extreme than the observed table.

-   For a two-sided test: $$
    p = \sum_{\text{tables with } P \le P_{\text{observed}}} P(\text{table})
    $$

-   Important properties:

    -   Exact test — no large-sample assumptions
    -   Valid even with very small counts
    -   Often conservative (p-values slightly larger than asymptotic
        tests)

------------------------------------------------------------------------

## Fisher’s Exact Test — What Does “More Extreme” Mean?

Suppose the margins are fixed as follows:

|       | Y = 1 | Y = 0 | Row Total |
|-------|-------|-------|-----------|
| X = 1 | a     | b     | 5         |
| X = 0 | c     | d     | 5         |
| Total | 4     | 6     | 10        |

So:

-   $n = 10$
-   $a+b = 5$
-   $c+d = 5$
-   $a+c = 4$

Under the null of independence, the only random quantity is $a$.

------------------------------------------------------------------------

### Step 1: Enumerate All Possible Tables

Since $a$ cannot exceed the margins:

$$
a \in \{0,1,2,3,4\}
$$

Each value of $a$ determines the entire table.

------------------------------------------------------------------------

### Compute Hypergeometric Probabilities (Exact Values) {.smaller}

Recall:

$$
P(a) =
\frac{\binom{5}{a}\binom{5}{4-a}}
{\binom{10}{4}}
$$

Compute for each possible $a$:

| $a$ | Interpretation | Probability |
|------------------------|------------------------|------------------------|
| 0 | all successes in X=0 | $\dfrac{\binom{5}{0}\binom{5}{4}}{\binom{10}{4}} =$ `r choose(5,0)*choose(5,4)/choose(10,4)` |
| 1 | mostly in X=0 | $\dfrac{\binom{5}{1}\binom{5}{3}}{\binom{10}{4}} =$ `r choose(5,1)*choose(5,3)/choose(10,4)` |
| 2 | roughly balanced | $\dfrac{\binom{5}{2}\binom{5}{2}}{\binom{10}{4}} =$ `r choose(5,2)*choose(5,2)/choose(10,4)` |
| 3 | mostly in X=1 | $\dfrac{\binom{5}{3}\binom{5}{1}}{\binom{10}{4}} =$ `r choose(5,3)*choose(5,1)/choose(10,4)` |
| 4 | all successes in X=1 | $\dfrac{\binom{5}{4}\binom{5}{0}}{\binom{10}{4}} =$ `r choose(5,4)*choose(5,0)/choose(10,4)` |

------------------------------------------------------------------------

### What Do We Notice?

-   The distribution is symmetric around $a=2$.
-   The most likely table is $a=2$ (balanced).
-   The least likely tables are $a=0$ and $a=4$ (most extreme).

If the observed table has $a=4$, the two-sided p-value is:

$$
P(a=4) + P(a=0)
$$

which equals:

`r choose(5,4)*choose(5,0)/choose(10,4) + choose(5,0)*choose(5,4)/choose(10,4)`

------------------------------------------------------------------------

### Step 3: Identify “More Extreme” Tables

Suppose the observed value is $a = 4$.

That corresponds to strong association (almost all successes in X=1).

For a two-sided test:

-   We include any table with probability $$
    P(\text{table}) \le P(a=4)
    $$

These will be the most imbalanced tables: - $a = 4$ - $a = 0$ (mirror
extreme)

------------------------------------------------------------------------

### Step 4: Compute the Two-Sided P-value

$$
p = P(a=4) + P(a=0)
$$

We are summing the probabilities of tables that are as rare or rarer
than the observed table.

------------------------------------------------------------------------

### Key Insight

“More extreme” does **not** mean:

-   Larger difference in proportions only.

It means:

> Smaller hypergeometric probability under the null.

Fisher’s exact test orders tables by their probability under $H_0$.

## Fisher’s Exact Test in `R`

In R:

```{r}
fisher.test(tab)
```

returns:

-   Exact p-value
-   Conditional maximum likelihood estimate of the odds ratio
-   Exact confidence interval

------------------------------------------------------------------------

## Final Summary

All methods test:

$$H_0: OR = 1$$

Differences arise from:

-   Asymptotic vs exact inference\
-   Wald vs score vs likelihood ratio approximations\
-   Continuity corrections

Next lecture: stratified tables and CMH → multivariable logistic
regression.

------------------------------------------------------------------------

## Compute Everything

```{r}
# Risks
p1_hat <- a/(a+b)
p0_hat <- c/(c+d)
RD_hat <- p1_hat - p0_hat

# sample sizes
n1 <- a + b
n0 <- c + d

se_rd <- sqrt( p1_hat*(1 - p1_hat)/n1 + p0_hat*(1 - p0_hat)/n0 )
z_crit <- qnorm(0.975)
RD_CI_lower <- RD_hat - z_crit * se_rd
RD_CI_upper <- RD_hat + z_crit * se_rd

# OR
OR_hat <- (a*d)/(b*c)

# Expected counts
E11 <- (a+b)*(a+c)/n
E12 <- (a+b)*(b+d)/n
E21 <- (c+d)*(a+c)/n
E22 <- (c+d)*(b+d)/n

# Pearson chi-square
chi2 <- (a-E11)^2/E11 + (b-E12)^2/E12 +
  (c-E21)^2/E21 + (d-E22)^2/E22
p_chi2 <- 1 - pchisq(chi2, df=1)

# Wald OR CI
logOR_hat <- log(OR_hat)
SE_logOR <- sqrt(1/a + 1/b + 1/c + 1/d)
Z_wald <- logOR_hat/SE_logOR
p_wald <- 2*(1 - pnorm(abs(Z_wald)))

CI_lower <- exp(logOR_hat - 1.96*SE_logOR)
CI_upper <- exp(logOR_hat + 1.96*SE_logOR)



list(
  RD_hat = RD_hat,
  RD_CI = c(RD_CI_lower, RD_CI_upper),
  OR_hat = OR_hat,
  OR_CI = c(CI_lower, CI_upper),
  chi2 = chi2,
  p_chi2 = p_chi2,
  Z_wald = Z_wald,
  p_wald = p_wald
)

```

------------------------------------------------------------------------

## Confirm With R Built-Ins

```{r}
prop.test(c(a,c), c(a+b, c+d), correct=FALSE)
prop.test(c(a,c), c(a+b, c+d), correct=TRUE)  

fisher.test(tab)

fit <- glm(survival ~ RHC, data=RHC, family=binomial)
summary(fit)
exp(confint(fit))

anova(fit,test = "Rao") ## Note p-value is equivalent to prop.test.
```
