---
title: "Survival Analysis — Comparing Survival Curves"
author: "Philip White"
format:
  revealjs:
    slide-number: true
    incremental: true
    theme: serif
    css: slides.css
    scrollable: true
    margin: 0
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(survminer)

theme_set(theme_minimal(base_size = 14))
```

------------------------------------------------------------------------

## Lecture 2 — Roadmap

1.  Review Kaplan–Meier curves
2.  Comparing survival curves visually
3.  Risk sets and expected deaths
4.  The **log-rank test**

------------------------------------------------------------------------

## Review: Kaplan–Meier Curves

Kaplan–Meier estimates survival as

$$
\widehat S(t) =
\prod_{t_j \le t}
\left(1 - \frac{d_j}{n_j}\right)
$$

where

-   $n_j$ = number at risk just before time $t_j$
-   $d_j$ = number of events at $t_j$

------------------------------------------------------------------------

## Example dataset - what are the event times?

::: fragment
-   `event = 1` → death occurred
-   `event = 0` → right censored
:::

::: fragment
| ID (we'll fix this soon) | Time | Event |
|:------------------------:|:----:|:-----:|
|            1             |  1   |   1   |
|            7             |  1   |   0   |
|            2             |  2   |   0   |
|            3             |  3   |   1   |
|            8             |  3   |   1   |
|            9             |  3   |   1   |
|            13            |  4   |   1   |
|            10            |  5   |   0   |
|            14            |  5   |   0   |
|            4             |  6   |   1   |
|            15            |  6   |   1   |
|            16            |  6   |   1   |
|            11            |  7   |   0   |
|            5             |  8   |   0   |
|            6             |  8   |   0   |
|            12            |  8   |   0   |
|            17            |  8   |   0   |
|            18            |  8   |   0   |
:::

------------------------------------------------------------------------

## Step-by-step KM calculation

::: fragment
### What to notice

-   The censored observation at time 2 lowers the risk set before time 3.
-   The two censored observations at time 5 lower the risk set before time 6.
-   KM only drops at event times, not censoring times.
:::

::: fragment
| Time | At Risk $n_j$ | Deaths $d_j$ | Conditional Survival $1 - d_j/n_j$ | Cumulative $S(t)$ |
|----|:--:|:--:|:--:|:--:|
| 0 | 18 | 0 | 1.0000 | 1.0000 |
| 1 | 18 | 1 | $1 - 1/18 = 17/18$ | $17/18 = 0.944$ |
| 3 | 15 | 3 | $1 - 3/15 = 12/15$ | $(17/18)(12/15) = 34/45 = 0.756$ |
| 4 | 12 | 1 | $1 - 1/12 = 11/12$ | $(34/45)(11/12) = 187/270 = 0.693$ |
| 6 | 9 | 3 | $1 - 3/9 = 6/9$ | $(187/270)(6/9) = 187/405 = 0.462$ |
:::

------------------------------------------------------------------------

## Overall KM Curve

::: panel-tabset
### Code

```{r,eval = FALSE}
# ----------------------------
# Create dataset and fit model
# ----------------------------
data_use <- data.frame(
  time = c(1,2,3,6,8,8,
           1,3,3,5,7,8,
           4,5,6,6,8,8),
  event = c(1,0,1,1,0,0,
            0,1,1,0,0,0,
            1,0,1,1,0,0),
  group = factor(c(rep("A",6), rep("B",6), rep("C",6)))
)


fit <- survfit(Surv(time, event) ~ 1, data = data_use)

# ----------------------------
# Plot KM curves
# ----------------------------
out <- suppressWarnings(
  ggsurvplot(
    fit,
    data = data_use,
    risk.table = TRUE,
    conf.int = FALSE,
    xlim = c(0, 8),
    linetype = "strata"
  )
)

out$plot
```

### Plot

```{r,echo = FALSE}

# ----------------------------
# Create dataset and fit model
# ----------------------------
data_use <- data.frame(
  time = c(1,2,3,6,8,8,
           1,3,3,5,7,8,
           4,5,6,6,8,8),
  event = c(1,0,1,1,0,0,
            0,1,1,0,0,0,
            1,0,1,1,0,0),
  group = factor(c(rep("A",6), rep("B",6), rep("C",6)))
)


fit <- survfit(Surv(time, event) ~ 1, data = data_use)

# ----------------------------
# Plot KM curves
# ----------------------------
out <- suppressWarnings(
  ggsurvplot(
    fit,
    data = data_use,
    risk.table = TRUE,
    conf.int = FALSE,
    xlim = c(0, 8),
    linetype = "strata"
  )
)
out$plot
```
:::

------------------------------------------------------------------------

## Comparing Survival Curves

::: fragment
Suppose we have **two or more** groups:

-   treatment vs. control
-   competing therapies
:::

::: fragment
We can estimate multiple KM curves. But the key question:

> Are the survival curves statistically different?
:::

------------------------------------------------------------------------

## Visual Comparison

-   Two curves may appear different.

-   But visual inspection can be misleading.

-   We need a **formal statistical test**.

------------------------------------------------------------------------

## Dataset with Group (18 observations)

::: fragment
-   `event = 1` → death occurred
-   `event = 0` → right censored
:::

::: fragment
| ID  | Group | Time | Event |
|:---:|:-----:|:----:|:-----:|
|  1  |   A   |  1   |   1   |
|  2  |   A   |  2   |   0   |
|  3  |   A   |  3   |   1   |
|  4  |   A   |  6   |   1   |
|  5  |   A   |  8   |   0   |
|  6  |   A   |  8   |   0   |
|  7  |   B   |  1   |   0   |
|  8  |   B   |  3   |   1   |
|  9  |   B   |  3   |   1   |
| 10  |   B   |  5   |   0   |
| 11  |   B   |  7   |   0   |
| 12  |   B   |  8   |   0   |
| 13  |   C   |  4   |   1   |
| 14  |   C   |  5   |   0   |
| 15  |   C   |  6   |   1   |
| 16  |   C   |  6   |   1   |
| 17  |   C   |  8   |   0   |
| 18  |   C   |  8   |   0   |
:::

------------------------------------------------------------------------

## Split the data by group

:::::: columns
::: {.column width="33%"}
### Group A

| ID  | Time | Event |
|:---:|:----:|:-----:|
|  1  |  1   |   1   |
|  2  |  2   |   0   |
|  3  |  3   |   1   |
|  4  |  6   |   1   |
|  5  |  8   |   0   |
|  6  |  8   |   0   |
:::

::: {.column width="33%"}
### Group B

| ID  | Time | Event |
|:---:|:----:|:-----:|
|  7  |  1   |   0   |
|  8  |  3   |   1   |
|  9  |  3   |   1   |
| 10  |  5   |   0   |
| 11  |  7   |   0   |
| 12  |  8   |   0   |
:::

::: {.column width="33%"}
### Group C

| ID  | Time | Event |
|:---:|:----:|:-----:|
| 13  |  4   |   1   |
| 14  |  5   |   0   |
| 15  |  6   |   1   |
| 16  |  6   |   1   |
| 17  |  8   |   0   |
| 18  |  8   |   0   |
:::
::::::

------------------------------------------------------------------------

## Group A — KM by hand

::: fragment
-   Event times for Group A: **1, 3, 6**
-   Censoring times for Group A: **2,8**
:::

::: fragment
### Step-by-step

| Time | At Risk $n_j$ | Deaths $d_j$ | Conditional Survival $1 - d_j/n_j$ | Cumulative $\widehat S(t)$ |
|:--:|:--:|:--:|:--:|:--:|
| 0 | 6 | 0 | 1 | 1 |
| 1 | 6 | 1 | $1 - 1/6 = 5/6$ | $1 \times 5/6 \approx 0.83$ |
| 3 | 4 | 1 | $1 - 1/4 = 3/4$ | $5/6 \times 3/4 = 5/8 \approx 0.625$ |
| 6 | 3 | 1 | $1 - 1/3 = 2/3$ | $5/8 \times 2/3 = 5/12 \approx 0.417$ |
:::

::: fragment
### What to notice

-   The censor at time 2 reduces the risk set before time 3.
-   The censors at time 8 do not change the curve, because there are no later event times.
:::

------------------------------------------------------------------------

## Group B — KM by hand

-   Event time for Group B: **3**
-   Censoring times for Group B: **1,5,7,8**

::: fragment
### Step-by-step

| Time | At Risk $n_j$ | Deaths $d_j$ | Conditional Survival $1 - d_j/n_j$ | Cumulative $S(t)$ |
|----|---:|---:|---:|---:|
| 0 | 6 | 0 | 1.0000 | 1.0000 |
| 3 | 5 | 2 | $1 - 2/5 = 3/5$ | $3/5 = 0.6$ |

### What to notice

-   The censor at time 1 removes one subject before the event time 3.
-   Both deaths at time 3 occur together, so the KM curve makes one drop there.
-   After time 3, there are no more deaths, so the curve stays flat.
:::

------------------------------------------------------------------------

## Group C — KM by hand

-   Event times for Group C: **4, 6**
-   Censoring times for Group C: \*1,5,7,8\*\*

### Step-by-step

| Time | At Risk $n_j$ | Deaths $d_j$ | Conditional Survival $1 - d_j/n_j$ | Cumulative $S(t)$ |
|----|---:|---:|---:|---:|
| 0 | 6 | 0 | 1.0000 | 1.0000 |
| 4 | 6 | 1 | $1 - 1/6 = 5/6$ | $5/6 = 0.83$ |
| 6 | 4 | 2 | $1 - 2/4 = 1/2$ | $5/12 = 0.417$ |

### What to notice

-   The censor at time 5 reduces the risk set before time 6.
-   Two deaths occur at time 6, so the curve drops once at that time using $d_j = 2$.

------------------------------------------------------------------------

## Combined survival values

Use this table to sketch the three KM curves on the same axes.

| Time | Group A | Group B | Group C |
|------|--------:|--------:|--------:|
| 0    |  1.0000 |  1.0000 |  1.0000 |
| 1    |  0.8333 |  1.0000 |  1.0000 |
| 2    |  0.8333 |  1.0000 |  1.0000 |
| 3    |  0.6250 |  0.6000 |  1.0000 |
| 4    |  0.6250 |  0.6000 |  0.8333 |
| 5    |  0.6250 |  0.6000 |  0.8333 |
| 6    |  0.4167 |  0.6000 |  0.4167 |
| 7    |  0.4167 |  0.6000 |  0.4167 |
| 8    |  0.4167 |  0.6000 |  0.4167 |

------------------------------------------------------------------------

## Plot in R

Which survival curve is best? Are they significantly different?

::: panel-tabset
### Code

```{r,eval = FALSE}
# ----------------------------
# Create dataset and fit model
# ----------------------------


fit <- survfit(Surv(time, event) ~ group, data = data_use)

# ----------------------------
# Plot KM curves
# ----------------------------
out <- suppressWarnings(
  ggsurvplot(
    fit,
    data = data_use,
    risk.table = TRUE,
    conf.int = FALSE,
    xlim = c(0, 8),
    linetype = "strata"
  )
)

out$plot
```

### Plot

```{r,echo = FALSE}
# ----------------------------
# Create dataset and fit model
# ----------------------------
data_use <- data.frame(
  time = c(1,2,3,6,8,8,
           1,3,3,5,7,8,
           4,5,6,6,8,8),
  event = c(1,0,1,1,0,0,
            0,1,1,0,0,0,
            1,0,1,1,0,0),
  group = factor(c(rep("A",6), rep("B",6), rep("C",6)))
)

fit <- survfit(Surv(time, event) ~ group, data = data_use)

# ----------------------------
# Plot KM curves
# ----------------------------
out <- suppressWarnings(
  ggsurvplot(
    fit,
    data = data_use,
    risk.table = TRUE,
    conf.int = FALSE,
    xlim = c(0, 8),
    linetype = "strata"
  )
)

out$plot
```
:::

------------------------------------------------------------------------

## Preparing for the Log-Rank Test

-   The log-rank test compares the survival curves between groups.
-   It is based on **observed** vs **expected** number of events in each group at each event time, summed across times.
-   We can now do the following
    -   compute the risk sets and events at each time (we did this),
    -   compute the expected counts under the null that survival is the same across groups (we actually did this too),
-   We will do the following
    -   Aggregate Observed vs. Expected to form the test statistic (we'll do a worked example next).

------------------------------------------------------------------------

## Key Idea Behind the Log-Rank Test

\begin{align*}
&H_0: S_A(t) = S_B(t) = S_C(t) = .... \\
&H_A: \text{At least one survival function is different.}
\end{align*}

At each event time we ask:

> Did more deaths occur in one group than expected?

We compare

-   **Observed deaths**
-   **Expected deaths under the null hypothesis**

------------------------------------------------------------------------

## Risk Set at an Event Time

At event time $t_j$:

|         | Group A  | Group B  | Group C  | Total |
|---------|----------|----------|----------|-------|
| At risk | $n_{Aj}$ | $n_{Bj}$ | $n_{Cj}$ | $n_j$ |
| Deaths  | $d_{Aj}$ | $d_{Bj}$ | $d_{Cj}$ | $d_j$ |

------------------------------------------------------------------------

## Expected Deaths

If survival is the same across groups, then expected deaths at event time $t_j$ in group $g$ is

$$
E_{gj} =
d_j
\frac{n_{gj}}{n_j}.
$$

-   Allocate deaths **proportionally to the risk set**
    -   $d_j$ - number of events across all groups at $t_j$.
    -   $n_{gj}$ - number in risk set of group $g$ at $t_j$
    -   $n_j$ - number in **total** risk set $g$ at $t_j$

------------------------------------------------------------------------

## Log-Rank Test Statistic w/ $k$ groups

For each group $g$ and time $t_j$:

$$
O_{gj} - E_{gj}
$$

Sum across all event times:

$$
\sum_j (O_{gj} - E_{gj})
$$

Test statistic:

$$
\chi^2 =
\sum_g \frac{(\sum_j (O_{gj} - E_{gj}))^2}{\text{Var}_g}
$$

Under $H_0$:

$$
\chi^2 \sim \chi^2_{k-1}
$$

-   Here: $k = 3$ groups → df = 2

------------------------------------------------------------------------

## Variance in the Log-Rank Test

::: fragment
At each event time $t_j$, the number of deaths in a group follows a **hypergeometric-like allocation**:

-   Total deaths: $d_j$
-   Total at risk: $n_j$
-   Group at risk: $n_{gj}$
:::

::: fragment
So the variance of $O_{gj}$ is:

$$
\mathrm{Var}(O_{gj}) =
d_j
\frac{n_{gj}}{n_j}
\left(1 - \frac{n_{gj}}{n_j}\right)
\frac{n_j - d_j}{n_j - 1}
$$
:::

------------------------------------------------------------------------

## Where Does This Come From?

This is the variance of a **hypergeometric distribution**:

-   We are distributing $d_j$ deaths across groups
-   Without replacement from the risk set

Interpretation:

-   First term: expected count
-   Second term: binomial variance component
-   Third term: finite population correction

------------------------------------------------------------------------

## Total Variance for Group $g$

We sum across event times:

$$
\mathrm{Var}_g =
\sum_j \mathrm{Var}(O_{gj})
$$

Then the log-rank statistic uses:

$$
\frac{\sum_j (O_{gj} - E_{gj})}
{\sqrt{\mathrm{Var}_g}}
$$

------------------------------------------------------------------------

## Key Insight

::: fragment
-   The log-rank test is essentially:

> A **sum of standardized deviations** between observed and expected deaths
:::

::: fragment
-   Same structure as:
    -   Chi-square tests
    -   Score tests
    -   Contingency tables
:::

------------------------------------------------------------------------

## Illustrative Example (Using Our Dataset)

Use the ordered dataset (by time):

| ID  | Group | Time | Event |
|:---:|:-----:|:----:|:-----:|
|  1  |   A   |  1   |   1   |
|  7  |   B   |  1   |   0   |
|  2  |   A   |  2   |   0   |
|  3  |   A   |  3   |   1   |
|  8  |   B   |  3   |   1   |
|  9  |   B   |  3   |   1   |
| 13  |   C   |  4   |   1   |
| 10  |   B   |  5   |   0   |
| 14  |   C   |  5   |   0   |
|  4  |   A   |  6   |   1   |
| 15  |   C   |  6   |   1   |
| 16  |   C   |  6   |   1   |
| 11  |   B   |  7   |   0   |
|  5  |   A   |  8   |   0   |
|  6  |   A   |  8   |   0   |
| 12  |   B   |  8   |   0   |
| 17  |   C   |  8   |   0   |
| 18  |   C   |  8   |   0   |

Event times: **1, 3, 4, 6**

------------------------------------------------------------------------

## Step 1 — Risk Sets

| Time | At Risk A | At Risk B | At Risk C | Total |
|------|-----------|-----------|-----------|-------|
| 1    | 6         | 6         | 6         | 18    |
| 3    | 4         | 5         | 6         | 15    |
| 4    | 3         | 3         | 6         | 12    |
| 6    | 2         | 2         | 5         | 9     |

------------------------------------------------------------------------

## Step 2 — Observed Deaths

| Time | $d_{Aj}$ | $d_{Bj}$ | $d_{Cj}$ | Total $d_j$ |
|------|----------|----------|----------|-------------|
| 1    | 1        | 0        | 0        | 1           |
| 3    | 1        | 2        | 0        | 3           |
| 4    | 0        | 0        | 1        | 1           |
| 6    | 1        | 0        | 2        | 3           |

------------------------------------------------------------------------

## Step 3 — Expected Deaths

Example at time 3:

Total deaths = 3\
Total at risk = 15

$$
E_A = 3 \times \frac{4}{15} = 0.8
$$

$$
E_B = 3 \times \frac{5}{15} = 1
$$

$$
E_C = 3 \times \frac{6}{15} = 1.2
$$

------------------------------------------------------------------------

## Step 4 — Contributions

Compute $O - E$ for each group at each time.

Example (Group A at time 3):

$$
1 - 0.8 = 0.2
$$

Interpretation:

-   Positive → more deaths than expected
-   Negative → fewer deaths than expected

------------------------------------------------------------------------

## Log-Rank Test in R

```{r}

output = with(data_use,survdiff(Surv(time,event) ~ group))

output$chisq
output$pvalue
```

------------------------------------------------------------------------

## Interpretation of the Log-Rank Test

-   The log-rank test compares **observed vs expected deaths**
-   It is a test of **whether survival curves differ across groups**
    -   It doesn't tell us *how* the curves differ.

::: fragment
> If survival is the same, observed ≈ expected at each time
:::

------------------------------------------------------------------------

## Assumptions of the Log-Rank Test

1.  Independent censoring
2.  Independent observations
3.  Hazards proportional over time (best power)

------------------------------------------------------------------------

## Connection to Cox Regression

The log-rank test is equivalent to

> the **score test from the Cox proportional hazards model**.

Next lecture:

-   Cox proportional hazards model
-   Hazard ratios
-   Regression interpretation

------------------------------------------------------------------------

## Signed (Directional) Log-Rank Statistic (Two Groups Only)

When there are **two groups** (e.g., Treatment vs Control), we can write a **signed test statistic**:

$$
Z =
\frac{\sum_j (O_{1j} - E_{1j})}
{\sqrt{\sum_j V_{1j}}}
\overset{\cdot}{\sim} N(0,1)
\quad \text{under } H_0
$$

-   Group 1 = treatment (for example)
-   Group 2 is determined automatically since totals must match

------------------------------------------------------------------------

## Interpretation of the Sign

The numerator:

$$
\sum_j (O_{1j} - E_{1j})
$$

tells us **direction of the effect**:

-   **Positive** $Z$:
    -   More deaths than expected in Group 1\
    -   → Worse survival in Group 1
-   **Negative** $Z$:
    -   Fewer deaths than expected in Group 1\
    -   → Better survival in Group 1

------------------------------------------------------------------------

## Why This Matters

-   The **signed statistic** lets you say:

> “Which group is doing better?”

-   The $\chi^2$ version only answers:

> “Are the groups different?”

------------------------------------------------------------------------

## Summary

Key ideas:

-   Survival curves can be compared formally
-   Log-rank compares **observed vs expected events**
-   Test statistic accumulates differences over time
-   R provides easy implementation with `survdiff()`
