---
title: "Count Outcomes: Poisson and Negative Binomial Regression"
subtitle: "Modeling Lyme Disease Counts by County"
author: "Philip White"
format:
  revealjs:
    slide-number: true
    incremental: true
    theme: serif
    css: slides.css
    scrollable: true
    margin: 0
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(MASS)        # glm.nb
library(broom)
library(patchwork)
library(usmap)
library(sf)
library(AER) ### dispersiontest

theme_set(theme_minimal(base_size = 14))

library(tidyverse)

state_temp <- tribble(
  ~stname,          ~mean_temp_F,
  "Alabama",        63.5,
  "Alaska",         28.1,
  "Arizona",        61.1,
  "Arkansas",       61.1,
  "California",     59.9,
  "Colorado",       46.4,
  "Connecticut",    49.9,
  "Delaware",       56.1,
  "Florida",        71.5,
  "Georgia",        64.3,
  "Hawaii",         70.2,
  "Idaho",          45.1,
  "Illinois",       52.7,
  "Indiana",        52.5,
  "Iowa",           48.7,
  "Kansas",         55.2,
  "Kentucky",       56.5,
  "Louisiana",      67.2,
  "Maine",          41.7,
  "Maryland",       55.4,
  "Massachusetts",  48.8,
  "Michigan",       45.3,
  "Minnesota",      42.2,
  "Mississippi",    64.2,
  "Missouri",       55.5,
  "Montana",        43.3,
  "Nebraska",       49.6,
  "Nevada",         51.1,
  "New Hampshire",  44.6,
  "New Jersey",     53.6,
  "New Mexico",     53.8,
  "New York",       46.1,
  "North Carolina", 59.9,
  "North Dakota",   41.3,
  "Ohio",           51.7,
  "Oklahoma",       60.6,
  "Oregon",         49.3,
  "Pennsylvania",   49.4,
  "Rhode Island",   50.9,
  "South Carolina", 63.3,
  "South Dakota",   46.1,
  "Tennessee",      58.5,
  "Texas",          65.3,
  "Utah",           49.5,
  "Vermont",        43.2,
  "Virginia",       56.3,
  "Washington",     49.1,
  "West Virginia",  52.7,
  "Wisconsin",      44.3,
  "Wyoming",        42.8
)
```

------------------------------------------------------------------------

## Learning goals

By the end of today:

-   Understand when count models are appropriate.
-   Derive and interpret the Poisson regression model.
-   Use offsets to model rates.
-   Diagnose overdispersion.
-   Fit and interpret Negative Binomial regression.
-   Compare Poisson vs NB models using LRT and AIC.

------------------------------------------------------------------------

## Remember the Lyme Dataset

```{r}
lyme <- read_csv("lyme_data.csv") %>% 
  mutate(
    case_rate_2022 = Cases2022 / TOT_POP,
    FIPS = paste0(
      sprintf("%02d", as.numeric(stcode)),
      sprintf("%03d", as.numeric(ctycode))
    )
  ) 

# County map with geometry
county_geo <- us_map("counties") %>%
  st_as_sf()

# Merge with Lyme data
lyme_geo <- county_geo %>%
  left_join(lyme, by = c("fips" = "FIPS")) %>%
  left_join(state_temp, by = "stname") %>% 
  mutate(
    centroid = st_centroid(geom),
    lat = st_coordinates(centroid)[,2],
    area_km2 = as.numeric(st_area(geom)) / 1e6,
    log_pop_density = log(TOT_POP / area_km2),
    case_rate_2022 = Cases2022 / TOT_POP,
    Northeast = stname %in% c("Massachusetts","Connecticut","New York",
                              "New Jersey","Pennsylvania","Rhode Island",
                              "Vermont","New Hampshire","Maine"
    )
  )  %>% 
  filter(
    !is.na(Cases2022),
    !is.na(TOT_POP),
    !is.na(mean_temp_F),
    !is.na(lat)
  )

glimpse(lyme_geo)


```

------------------------------------------------------------------------

## Mapping 2022 annual incidence rate

```{r}

usa_data <- us_map("counties") %>% 
  left_join(lyme, by = c("fips" = "FIPS"))

usa_states <- us_map("states")

ggplot(usa_data) + 
  theme_bw() +
  geom_sf(aes(fill = case_rate_2022)) +
  geom_sf(data = usa_states, fill = NA, col = "black", linewidth = 0.35) +
  scale_fill_gradientn(
    colors = c("white", "orange", "magenta", "darkred"),
    labels = scales::label_percent()
  ) +
  labs(
    fill = "Annual\nIncidence\nRate",
    title = "Lyme Disease per-year Incidence Rate (2022)"
  )
  
```

------------------------------------------------------------------------

## Final Map

```{r,echo = FALSE}
ggplot(usa_data) + 
  theme_bw() +
  geom_sf(aes(fill = case_rate_2022)) +
  geom_sf(data = usa_states, fill = NA, col = "black", linewidth = 0.35) +
  scale_fill_gradientn(
    colors = c("white", "orange", "magenta", "darkred"),
    labels = scales::label_percent()
  ) +
  labs(
    fill = "Annual\nIncidence\nRate",
    title = "Lyme Disease per-year Incidence Rate (2022)"
  )
  
```

------------------------------------------------------------------------

## Final Map - Counts

```{r,echo = FALSE}
ggplot(usa_data) + 
  theme_bw() +
  geom_sf(aes(fill = Cases2022)) +
  geom_sf(data = usa_states, fill = NA, col = "black", linewidth = 0.35) +
  scale_fill_gradientn(
    colors = c("white", "orange", "magenta", "darkred"),
    # labels = scales::label_percent()
  ) +
  labs(
    fill = "Annual\nCounts",
    title = "Lyme Disease per-year Incidence Rate (2022)"
  )
  
```

------------------------------------------------------------------------

## Relationships between rate and temperature

```{r,echo = FALSE}
ggplot(lyme_geo) + 
  theme_bw(base_size = 16) +
  geom_point(aes(x = mean_temp_F,y = case_rate_2022,col = Northeast))+
    scale_y_continuous(
    labels = scales::label_percent()
  ) +
  labs(
    x = "Mean Temperature (F)",
    y = "Lyme Incidence Rate"
  )

```

## Relationships between Population Density and Rate

```{r,echo = FALSE}

ggplot(lyme_geo) + 
  theme_bw(base_size = 16) +
  geom_point(aes(x = log_pop_density,y = case_rate_2022,col = Northeast))+
    scale_y_continuous(
    labels = scales::label_percent()
  ) +
  labs(
    x = "Log Population Density",
    y = "Lyme Incidence Rate"
  )

```

## Relationships between State Status and Rate

```{r,echo = FALSE}

ggplot(lyme_geo %>% filter(!is.na(stname))) + 
  theme_bw(base_size = 16) +
  geom_boxplot(aes(x = ststatus,y = case_rate_2022))+
    scale_y_continuous(
    labels = scales::label_percent()
  ) +
  theme(
  ) + 
  labs(
    x = "State Status",
    y = "Incidence Rate"
  )
  
```

## Relationships between State and Rate

```{r,echo = FALSE}

ggplot(lyme_geo %>% filter(!is.na(stname))) + 
  theme_bw(base_size = 16) +
  geom_boxplot(aes(x = stname,y = case_rate_2022))+
    scale_y_continuous(
    labels = scales::label_percent()
  ) +
  theme(
    axis.text.x = element_text(angle = 270,hjust= 0,vjust= .5)
  ) + 
  labs(
    x = "",
    y = "Incidence Rate"
  )
  
```

------------------------------------------------------------------------

## Distribution of Counts

```{r}
ggplot(lyme, aes(x = Cases2022)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  labs(title = "Distribution of Lyme Cases (2022)")
```

Discussion:

-   Is this symmetric?
-   Is variance constant?
-   What does this suggest?

------------------------------------------------------------------------

## Why Not Linear Regression?

Lyme disease cases per county are:

-   Non-negative integers.
-   Highly skewed.
-   Variance often increases with the mean.
-   Many counties have zero cases.

OLS assumptions are violated.

We will model:

-   Outcome: `Cases2022`
-   Exposure: `TOT_POP`, Geography (`stname`,`Northeast`),
    `log_pop_density`, `mean_temp_F`, `ststatus`

------------------------------------------------------------------------

## The Poisson Distribution

If $Y \sim \text{Poisson}(\mu)$:

-   $E(Y) = \mu$

-   $\text{Var}(Y) = \mu$

-   Key property:

> Mean equals variance.

This is often unrealistic in real data.

------------------------------------------------------------------------

## Poisson Regression Model

Like logistic regression, this is another example of a generalized
linear model

Assume:

$$
\text{Cases}_i \overset{ind}{\sim} \text{Poisson}(\mu_i)
$$

Log link:

$$
\log(\mu_i) = \beta_0 + X_i^\top \beta
$$

Interpretation:

-   $\beta_j$ = change in the log of the expected count, holding all
    other variables constant.
-   $e^{\beta_j}$ = multiplicative change in expected count, holding all
    other variables constant.

------------------------------------------------------------------------

## Correct Distributional Form

-   The outcome is a **count** (non-negative integers).

-   Conditional on covariates:

    $$
    E(Y_i \mid X_i) = \mu_i
    $$

    $$
    \mathrm{Var}(Y_i \mid X_i) = \mu_i
    $$

-   Mean equals variance (**equidispersion**).

------------------------------------------------------------------------

## Log-Linear Mean Structure

-   The log of the expected count is linear in predictors: $$
    \log(\mu_i) = X_i^\top \beta
    $$

-   Implies multiplicative effects on the count scale: $$
    e^{\beta_j} = \text{Incidence Rate Ratio (IRR)}
    $$

------------------------------------------------------------------------

## Independence

-   Observations are independent conditional on covariates.
-   No unmodeled clustering or spatial dependence.

------------------------------------------------------------------------

## Consequences of Violations

-   Overdispersion → underestimated standard errors.
-   Wrong functional form → biased IRRs.
-   Omitted variables.

If assumptions fail, we could consider alternative models:

-   Negative Binomial
-   Zero-inflated models
-   Quasi-Poisson models
-   Spatial/time-series models

------------------------------------------------------------------------

## Modeling Rates with Offsets

To model rates, we would typically use an **offset** for population:

We model:

$$
\log(\mu_i) = \beta_0 \log(\text{Population}_i) + X_i^\top \beta
$$

Equivalent to modeling:

$$
\log\left(\frac{\mu_i}{\text{Population}_i}\right) = X_i^\top \beta
$$

The term $\log(\text{Population}_i)$ is an **offset**. It is a covariate
w/o a regression coefficient.

------------------------------------------------------------------------

## Fit Poisson Model (Counts)

```{r}
fit_pois <- glm(Cases2022 ~ TOT_POP,
                data = lyme,
                family = poisson)

summary(fit_pois)
```

This is wrong because population should be an offset.

------------------------------------------------------------------------

## Fit Poisson Model with Offset

```{r}
fit_pois_offset <- glm(Cases2022 ~ ststatus + mean_temp_F + log_pop_density,
                       offset = log(TOT_POP),
                       data = lyme_geo,
                       family = poisson)

summary(fit_pois_offset)
```

Interpretation:

-   Intercept represents baseline log incidence rate.

------------------------------------------------------------------------

## Add Predictors

Suppose we use:

-   State status (`ststatus`)
-   State-level average temperature (`mean_temp_F`)
-   Log population density (`log_pop_density`)

```{r}
fit_pois2 <- glm(Cases2022 ~ ststatus + mean_temp_F + log_pop_density,
                 offset = log(TOT_POP),
                 data = lyme_geo,
                 family = poisson)

summary(fit_pois2)
```

Interpretation: - $\beta_{temp}$ = change in the log-mean for a one unit
change in avg. temperature, holding other variables constant. -
$e^{\beta}$ = incidence rate ratio (IRR) for a one unit change in avg.
temperature, holding other variables constant.

------------------------------------------------------------------------

## Interpretation of IRR for `ststatus`

```{r}
summary(fit_pois2)
```

If $\hat{\beta} = -3.9814524$:

$$
e^{-3.9814524} \approx 0.019
$$

Interpretation:

-   Counties in "Low Incidence" states have 98% fewer cases, on average,
    than cases in "High Incidence" states, holding other variables
    constant.

------------------------------------------------------------------------

## What Does the Poisson Model Assume?

If

$$
Y_i \sim \text{Poisson}(\mu_i),
$$

then the Poisson distribution implies:

$$
E(Y_i) = \mu_i
$$

$$
\mathrm{Var}(Y_i) = \mu_i
$$

**Key assumption:**

> The variance equals the mean.

This is called **equidispersion**.

------------------------------------------------------------------------

## Why Might This Be Unrealistic?

Real epidemiologic count data often show:

-   Geographical patterns
-   Unmeasured heterogeneity
-   Outbreak behavior
-   Zero inflation (more zeros than a Poisson would expect)

These can create:

$$
\mathrm{Var}(Y_i) > E(Y_i)
$$

This is called **overdispersion**.

If we ignore it:

-   Standard errors are too small
-   p-values are too optimistic
-   Confidence intervals are too narrow

------------------------------------------------------------------------

## Checking for Overdispersion

A simple diagnostic for Poisson GLMs:

$$
\frac{\text{Residual Deviance}}{\text{Residual Degrees of Freedom}}
$$

In R:

```{r}
deviance(fit_pois2) / df.residual(fit_pois2)
```

Interpretation:

-   ≈ 1 → Poisson variance assumption reasonable

-   

    > > 1 → Overdispersion present

-   \<\< 1 → Underdispersion (rare in practice)

If overdispersion is present:

> The Poisson assumption is violated → consider Negative Binomial.

------------------------------------------------------------------------

## Another Quick Check

```{r}
mean(lyme$Cases2022)
var(lyme$Cases2022)
```

If variance \>\> mean → overdispersion likely.

------------------------------------------------------------------------

## More formal test

```{r}
### AER package
dispersiontest(fit_pois2)
```

------------------------------------------------------------------------

$$
Y_i \sim \text{Negative Binomial}(\mu_i, \theta),
$$

with

$$
E(Y_i \mid X_i) = \mu_i
$$

$$
\mathrm{Var}(Y_i \mid X_i) = \mu_i + \frac{\mu_i^2}{\theta}.
$$

------------------------------------------------------------------------

### Mean Model (same as Poisson)

We use the same log link:

$$
\log(\mu_i) = X_i^\top \beta
$$

or, when modeling rates,

$$
\log(\mu_i) = \log(\text{Exposure}_i) + X_i^\top \beta.
$$

So interpretation of coefficients is unchanged:

$$
e^{\beta_j} = \text{Incidence Rate Ratio (IRR)}.
$$

------------------------------------------------------------------------

## Role of the Dispersion Parameter

-   $\theta$ controls overdispersion.

-   As $\theta \to \infty$: $$
    \mathrm{Var}(Y_i) \to \mu_i
    $$ and the model approaches Poisson.

-   Small $\theta$ → large overdispersion.

------------------------------------------------------------------------

## Intuition

The Negative Binomial can be viewed as:

> A Poisson model with unobserved heterogeneity in the rate.

It allows

$$
\mathrm{Var}(Y_i) > E(Y_i),
$$

which is common in epidemiologic count data.

------------------------------------------------------------------------

## Fit Negative Binomial

```{r}
fit_nb <- glm.nb(Cases2022 ~ ststatus + mean_temp_F + log_pop_density + offset(log(TOT_POP)),
                 data = lyme_geo)

summary(fit_nb)
```

Interpretation of coefficients same as Poisson:

-   $e^{\beta}$ = incidence rate ratio.

------------------------------------------------------------------------

## Compare Models

```{r}
AIC(fit_pois2, fit_nb)
```

Lower AIC preferred.

------------------------------------------------------------------------

## 16. Likelihood Ratio Test

```{r}
anova(fit_pois2, fit_nb, test = "LRT")
```

If NB significantly better → prefer NB.

------------------------------------------------------------------------

## 17. Predicted Rates

```{r}
lyme_geo$pred_nb <- predict(fit_nb, type = "response")

ggplot(lyme_geo, aes(x = TOT_POP, y = pred_nb)) +
  geom_point(alpha = 0.4) +
  labs(title = "Predicted Counts (Negative Binomial)")
```

------------------------------------------------------------------------

## Conceptual Summary

Poisson:

-   Mean = Variance.
-   Good when counts small, dispersion minimal.

Negative Binomial:

-   Handles overdispersion.
-   More flexible.
-   Often preferred in epidemiology.

------------------------------------------------------------------------

## In-Class Modeling Exercise

In groups:

1.  Fit Poisson with offset.
2.  Check overdispersion.
3.  Fit NB model.
4.  Compare AIC.
5.  Interpret one IRR in plain language.
6.  Discuss: which model is better and why?

------------------------------------------------------------------------

## Discussion Prompts

-   Why does modeling counts require a log link?
-   When is offset necessary?
-   When might zero-inflated models be appropriate?
-   How does interpretation compare to logistic regression?

------------------------------------------------------------------------

## Takeaways

-   Count models are extensions of GLMs.
-   Offsets allow modeling of rates.
-   Overdispersion matters.
-   Negative Binomial is often more realistic.
-   Interpretation always occurs on multiplicative scale.
