---
title: "Cox Proportional Hazards Model"
author: "Philip White"
format:
  revealjs:
    slide-number: true
    incremental: true
    theme: serif
    css: slides.css
    scrollable: true
    margin: 0
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(survminer)
theme_set(theme_minimal(base_size = 14))
```

------------------------------------------------------------------------

## Roadmap

1.  From comparing curves → modeling covariates\
2.  The Cox proportional hazards model\
3.  Interpretation of coefficients\
4.  Why no baseline hazard or intercept?\
5.  Likelihood (conceptual)\
6.  Example with data

------------------------------------------------------------------------

## From KM Curves to Regression

::: fragment
So far, we have:

-   Estimated survival curves (Kaplan–Meier)
-   Compared curves across groups (log-rank test)
:::

::: fragment
But what if we want to:

-   Adjust for multiple variables?
-   Model continuous predictors?
-   Estimate effects simultaneously?
:::

::: fragment
> We need a **regression model for survival data**.
:::

------------------------------------------------------------------------

## Goal: Model the Hazard

Instead of modeling survival directly, we model:

$$
h(t \mid X)
$$

::: fragment
> The hazard: instantaneous risk at time $t$ given survival up to $t$.

$$
h(t | X) =
\lim_{\Delta t \to 0}
\frac{P(t < T \le t + \Delta t \mid T \ge t, X)}{\Delta t}
$$

Equivalent expression:

$$
h(t | X) = \frac{f(t | X)}{S(t | X)}
$$
:::

::: fragment
We want to understand how covariates affect this hazard.
:::

------------------------------------------------------------------------

## Cox Proportional Hazards Model

The Cox model assumes:

$$
h(t \mid X) = h_0(t)\exp(X^\top \beta)
$$

::: fragment
### Breaking Down the Model

-   $h(t \mid X)$: hazard for an individual with covariates $X$
:::

::: fragment
-   $h_0(t)$: **baseline hazard**
    -   hazard when $X = 0$
    -   This may not be interpretable.
:::

::: fragment
-   $\exp(X^\top \beta)$: multiplicative effect of covariates
:::

------------------------------------------------------------------------

## Key Interpretation

Compare two individuals with covariates $X_1$ and $X_2$:

$$
\frac{h(t \mid X_1)}{h(t \mid X_2)} =
\exp\big((X_1 - X_2)^\top \beta\big)
$$

### What This Means

::: fragment
-   The model describes **relative risk**
:::

::: fragment
-   The ratio **does not depend** on time or the baseline hazard
:::

::: fragment
> This is the **proportional hazards assumption**
:::

------------------------------------------------------------------------

## Interpreting Coefficients

Focusing on one covariate $X_j$,

-   $\exp(\beta_j)$ is the *true* **hazard ratio** associated with
    -   a one unit increase in $X_j$,
    -   holding other covariates constant.

::: fragment
### Examples

If $\exp(\beta_j) = 1.5$ $\rightarrow$ 50% higher hazard associated with a one unit increase in $X_j$, holding all other covariates fixed.
:::

::: fragment
If $\exp(\beta_j) = 0.7$ $\rightarrow$ 30% lower hazard associated with a one unit increase in $X_j$, accounting for other model covariates.
:::

::: fragment
> Interpretation is always in terms of **hazard (risk), not probability**\
> A hazard ratio compares **risk over time**, not survival probability directly
:::

------------------------------------------------------------------------

## What Is Assumed?

::: fragment
1.  Independent observations
:::

::: fragment
2.  Independent (non-informative) censoring
:::

::: fragment
3.  Correct specification of covariates
:::

::: fragment
**4. Proportional Hazards**
:::

------------------------------------------------------------------------

## Key Assumption: Proportional Hazards

The key assumption is:

$$
\frac{h(t \mid X_1)}{h(t \mid X_2)} =
\exp\big((X_1 - X_2)^\top \beta\big)
$$

::: fragment
The hazard ratio is **constant over time**.
:::

::: fragment
-   If one group has higher risk, it stays higher at all times
:::

::: fragment
-   The curves may separate, but they do not cross (in theory)
:::

::: fragment
> This is why the model is called **proportional hazards**
:::

------------------------------------------------------------------------

## What Is NOT Assumed?

::: fragment
We do **not** assume a specific distribution for survival times.
:::

::: fragment
-   Not exponential
-   Not Weibull
-   Not normal
:::

::: fragment
> The Cox model is **semi-parametric**
:::

------------------------------------------------------------------------

## Why Is It Called Semi-Parametric?

::: fragment
The model has two parts:

$$
h(t \mid X) = h_0(t)\exp(X^\top \beta)
$$
:::

::: fragment
-   $h_0(t)$: Will not estimate (nonparametric)
-   We do not assume a specific distributional form for the hazard.
-   $\beta$: Will estimate (parametric)
:::

------------------------------------------------------------------------

## Why No Intercept?

Suppose we write:

$$
h(t \mid X) = h_0(t)\exp(\beta_0 + X^\top \beta)
$$

::: fragment
We can rewrite this as:

$$
h(t \mid X) =
\underbrace{h_0(t)\exp(\beta_0)}_{\text{new baseline hazard}}
\exp(X^\top \beta)
$$
:::

::: fragment
So:

-   $\beta_0$ gets absorbed into $h_0(t)$
:::

::: fragment
> The intercept is not identifiable
:::

------------------------------------------------------------------------

## Why Don’t We Estimate $h_0(t)$?

Look at the hazard ratio:

$$
\frac{h(t \mid X_1)}{h(t \mid X_2)}
=
\frac{h_0(t)\exp(X_1^\top \beta)}
     {h_0(t)\exp(X_2^\top \beta)}
$$

::: fragment
The baseline hazard cancels:

$$
=
\exp\big((X_1 - X_2)^\top \beta\big)
$$
:::

::: fragment
> The model depends only on **relative risk**\
> We never need to estimate $h_0(t)$ to estimate $\beta$
:::

::: fragment
This is what makes the Cox model so powerful

-   No distributional assumptions
-   Still get interpretable coefficients
:::

------------------------------------------------------------------------

## From KM to Cox Partial Likelihood

::: fragment
Recall:

-   Kaplan–Meier: uses **risk sets**
-   Log-rank: compares **observed vs expected deaths**
:::

::: fragment
Cox model focuses on the following question:

> Given that an event occurs at $t_j$, **who in the risk set has the event?**
:::

::: fragment
This idea leads directly to the *partial* likelihood (Cox's Idea)
:::

------------------------------------------------------------------------

## What is a Partial Likelihood? (Cox’s Idea)

::: fragment
The full likelihood depends on:

$$
L(\beta, h_0(\cdot))
$$

-   regression parameters $\beta$
-   baseline hazard $h_0(t)$ (unknown and not of interest)
:::

::: fragment
Cox’s key idea:

> Separate the likelihood into two pieces

$$
L(\beta, h_0(\cdot)) = L_1(\beta) \cdot L_2(\beta, h_0)
$$
:::

::: fragment
-   $L_1(\beta)$ contains **most of the information about** $\beta$
-   $L_2(\beta, h_0)$ depends heavily on the unknown baseline hazard
:::

------------------------------------------------------------------------

## Partial Likelihood

::: fragment
We focus only on:

$$
L_1(\beta)
$$
:::

::: fragment
### Key Properties

-   Uses only information at **event times**
-   Based on **who fails within each risk set**
-   Does **not require modeling** $h_0(t)$
:::

::: fragment
### Important Point

> Partial likelihood is used **only to estimate** $\beta$

-   It is not a full likelihood for the entire survival distribution
-   But it is sufficient for estimating **hazard ratios**
:::

------------------------------------------------------------------------

## Intuition of Cox Model

::: fragment
To recap, Cox model focuses on the following question:

> Given that an event occurs at $t_j$, **who in the risk set has the event?**
:::

::: fragment
Multiply those probabilities across all event times:

→ this gives the **partial likelihood**
:::

::: fragment
This idea leads directly to the *partial* likelihood (Cox's key idea)
:::

------------------------------------------------------------------------

## Risk Set

At every event time $t_j$, let

::: fragment
$$
n_j = \text{set of individuals at risk at time } t_j
$$
:::

::: fragment
This is the same risk set used in:

-   Kaplan–Meier
-   Log-rank test
:::

::: fragment
For simplicity, we are going to focus on the scenario where all event times are unique.

-   That is, there are no ties between
:::

------------------------------------------------------------------------

## Probability of an Event

Suppose an individual $i$ dies at time $t_j$.

What is the probability that *this individual* dies at that time?

::: fragment
$$
P(i \text{ dies at } t_j)=\frac{h(t_j \mid X_i)}{\sum_{k \in n_j} h(t_j \mid X_k)}
$$
:::

------------------------------------------------------------------------

## Plug in the Cox Model

Recall:

$$
h(t \mid X) = h_0(t)\exp(X^\top \beta)
$$

::: fragment
Substitute into the probability:

$$
P(i \text{ dies at } t_j)=\frac{h(t_j \mid X_i)}{\sum_{k \in n_j} h(t_j \mid X_k)}=
\frac{h_0(t_j)\exp(X_i^\top \beta)}
{\sum_{k \in n_j} h_0(t_j)\exp(X_k^\top \beta)}
$$
:::

::: fragment
The baseline hazard cancels:

$$
P(i \text{ dies at } t_j)=
\frac{\exp(X_i^\top \beta)}
{\sum_{k \in n_j} \exp(X_k^\top \beta)}
$$
:::

::: fragment
> This is why we do **not need to estimate** $h_0(t)$
:::

------------------------------------------------------------------------

## Partial Likelihood

We multiply these probabilities across all event times:

$$
L(\beta) =
\prod_j
\frac{\exp(X_{(j)}^\top \beta)}
{\sum_{k \in n_j} \exp(X_k^\top \beta)}
$$

::: fragment
### What This Means

Each term represents:

> Probability that the observed individual fails at that time
:::

::: fragment
Given:

-   everyone who was at risk
:::

------------------------------------------------------------------------

## How Does Censoring Enter the Partial Likelihood?

::: fragment
Recall:

$$
L(\beta) =
\prod_j
\frac{\exp(X_{(j)}^\top \beta)}
{\sum_{k \in n_j} \exp(X_k^\top \beta)}
$$
:::

::: fragment
Only **event times** contribute terms to the likelihood.
:::

::: fragment
Censored observations:

-   do **not** create a term in the likelihood\
-   do **not** appear in the numerator
:::

------------------------------------------------------------------------

### Where Do Censored Observations Matter?

::: fragment
They affect the **risk set**:

$$
n_j = \text{individuals still under observation just before } t_j
$$
:::

::: fragment
-   If someone is censored before $t_j$, they are **removed from** $n_j$
-   If censored after $t_j$, they are still included
:::

------------------------------------------------------------------------

### Why This Works

::: fragment
At each event time, we are asking:

> “Who fails among those still at risk?”
:::

::: fragment
Censored individuals:

-   contribute information **up to their censoring time**
-   then leave the risk set
:::

::: fragment
-   Events → contribute to the likelihood\
-   Censoring → changes the risk set
:::

::: fragment
> This is the same idea used in Kaplan–Meier and log-rank
:::

------------------------------------------------------------------------

## Connection to Log-Rank Test

::: fragment
Log-rank test:

-   compares observed vs expected deaths
:::

::: fragment
Cox model:

-   models probability of who dies at each time
:::

::: fragment
> The log-rank test is the **score test from the Cox model**\
> when the model includes **only a group indicator**
:::

::: fragment
With additional covariates:

-   the Cox model generalizes the log-rank test
-   allowing adjustment for multiple variables
:::

------------------------------------------------------------------------

## Recap

::: fragment
The Cox model asks:

> “Given the risk set, who is most likely to fail?”
:::

::: fragment
This depends only on:

-   $\exp(X^\top \beta)$
:::

::: fragment
Not on:

-   baseline hazard
:::

------------------------------------------------------------------------

## Why This Is Powerful

::: fragment
-   No need to model time distribution
-   Works with censored data
-   Handles multiple covariates
:::

::: fragment
> We get interpretable regression without specifying $h_0(t)$
:::

------------------------------------------------------------------------

## Comparison of methods

::: fragment
KM → survival estimation

Log-rank → group comparison

Cox → regression model
:::

::: fragment
All three use:

> **risk sets over time**
:::

------------------------------------------------------------------------

## Example: Cox Model with Lung Data

We now apply the Cox model to real data.

The `lung` dataset contains:

-   `time`: survival time (days)
-   `status`: 1 = censored, 2 = dead
-   `ph.ecog`: ECOG performance score rated by the physician

### ECOG performance score

-   `0` = asymptomatic
-   `1` = symptomatic but completely ambulatory
-   `2` = in bed less than 50% of the day
-   `3` = in bed more than 50% of the day but not bedbound
-   `4` = bedbound (not present in data)

------------------------------------------------------------------------

## Data Preparation

```{r}
lung_use <- lung %>%
  mutate(
    ph.ecog = case_when(
      ph.ecog %in% c(2, 3) ~ "2-3",
      ph.ecog == 0 ~ "0",
      ph.ecog == 1 ~ "1",
      ph.ecog == 4 ~ "4",
      TRUE ~ NA_character_
    ),    
    event = ifelse(status == 1, 0, 1),
    sex = ifelse(sex == 1, "Male","Female")
  ) %>%
  filter(!is.na(time), !is.na(event), !is.na(ph.ecog))
```

------------------------------------------------------------------------

## Cox Model (Single Statification Factor)

```{r}
fit1 <- coxph(Surv(time, event) ~ ph.ecog, data = lung_use)
summary(fit1)
```

------------------------------------------------------------------------

## Interpreting Output

::: fragment
Focus on:

-   $\beta$ → change in log-hazard ratio.
-   $\exp(\beta)$ → hazard ratio
:::

::: fragment
In our example, each coefficient compares a group to the `ph.ecog = 0` **reference level**.
:::

------------------------------------------------------------------------

## Interpretation Example

::: fragment
Suppose:

$$
\exp(\beta) \approx 1.45
$$
:::

::: fragment
Interpretation:

> Individuals w/ `ph.ecog = 1` have **45% higher hazard** than `ph.ecog = 0`.
:::

::: fragment
That means:

-   Higher instantaneous risk of death
-   Lower survival over time
:::

------------------------------------------------------------------------

## Your Turn (2-3 minutes to answer these)

::: fragment
-   Interpret $\exp(\widehat\beta)$ for `ph.ecog2-3`
-   What if I want to know the effect of having `ph.ecog = 2-3`, relative to `ph.ecog = 1` (not the reference group)?
    -   What should/can we do?
:::

```{r}
summary(fit1)
```

------------------------------------------------------------------------

## Important: Hazard vs Survival

::: fragment
Hazard ratio:

-   compares **instantaneous risk**
:::

::: fragment
It does **not directly give survival probabilities**
:::

::: fragment
But:

> Higher hazard → lower survival
:::

------------------------------------------------------------------------

## Add More Covariates

```{r}
fit2 <- coxph(Surv(time, event) ~ ph.ecog + age + sex, data = lung_use)
summary(fit2)
```

::: fragment
-   Interpret $\exp(\widehat\beta)$ for `sex` with your neighbor
-   How would you explain this to a non-statistician?
:::

------------------------------------------------------------------------

## Interpreting Continuous Variables

::: fragment
For age:

$$
\exp(\widehat \beta_{\text{age}}) \approx 1.01
$$
:::

::: fragment
Interpretation:

> Each 1-year increase in age multiplies the hazard by 1.01, holding sex and physician ECOG fixed. → 1% increase in hazard per year
:::

------------------------------------------------------------------------

## Compare Two Individuals

::: fragment
Suppose:

-   Person A: age 70\
-   Person B: age 60
:::

::: fragment
Then:

$$
\frac{h_A(t)}{h_B(t)} =
\exp(10 \cdot \widehat\beta_{\text{age}}) \approx 1.12
$$
:::

::: fragment
Interpretation:

> A 10-year age difference increases risk by 12%, holding all other variables fixed. That is, it compounds the hazard ratio.
:::

------------------------------------------------------------------------

## Connection to KM Curves

::: fragment
In the homework, you will examine and test:

-   KM curves for `ph.ecog`
:::

::: fragment
Now:

-   Cox model quantifies those differences
:::

::: fragment
> KM shows patterns\
> Cox explains them
:::

------------------------------------------------------------------------

## As you do the homework

::: fragment
Ask:

-   Which group had the worst survival?
:::

::: fragment
Check:

-   Does that group have the largest hazard ratio?
:::

::: fragment
> This consistency is important
:::

------------------------------------------------------------------------

## Interpretation Practice w/ more complex model

::: fragment
Pick one continuous (age, wt.loss in last 6 months) and one categorical (ph.ecog or sex) coefficient from the output.
:::

::::::: columns
:::: {.column width="50%"}
::: fragment
Write:

-   The hazard ratio
-   A full sentence interpretation
:::
::::

:::: {.column width="50%"}
::: fragment
Be precise:

> “Holding other variables constant…”
:::
::::
:::::::

```{r}
fit3 <- coxph(Surv(time, event) ~ ph.ecog + age + sex +  wt.loss, data = lung_use)
summary(fit3)
```

------------------------------------------------------------------------

## Schoenfeld Residuals

::: fragment
Schoenfeld residuals are used to check the **proportional hazards assumption**.

At each event time, they measure:

> How different the observed covariate value is from what the model expects
:::

::: fragment
If the proportional hazards assumption holds:

-   Residuals should show **no trend over time**
-   They should fluctuate randomly around 0
:::

::: fragment
### What would be a problem?

-   Systematic upward or downward trend
-   Clear pattern over time

> This suggests the effect of a covariate is **changing over time**
:::

------------------------------------------------------------------------

## What Are Schoenfeld Residual Plots Showing?

::: fragment
At each event time:

-   Compare an individual's covariate value\
-   To the **average covariate value in the risk set**
:::

::: fragment
This difference is a **Schoenfeld residual**:

> observed − expected (based on the risk set)
:::

::: fragment
### Scaled Schoenfeld Residuals

Residuals are scaled by their variability

-   downweight noisy (high-variance) values\
-   standardize across time
:::

------------------------------------------------------------------------

## We'll look at plots over time

::: fragment
The plot shows:

> An estimate of how the coefficient $\beta_j$ changes over time
:::

::: fragment
Formally:

$$
\beta_j(t) \approx \widehat{\beta}_j + \text{scaled residual}
$$
:::

### How to Interpret the Plot

::: fragment
-   Flat line → effect is constant → PH assumption holds
:::

::: fragment
-   Trend over time → effect changes → PH violation
:::

::: fragment
Example:

-   Strong early effect → weak later effect\
-   → hazard ratio is not constant
:::

------------------------------------------------------------------------

## Testing the PH Assumption

We can formally test this using `cox.zph()`:

```{r}
zph3 <- cox.zph(fit3)
zph3
```

::: fragment
-   Tests whether residuals are correlated with time
-   Provides:
    -   a test for each covariate
    -   a **global test** for the model
-   Large p-value (e.g., \> 0.05):
    -   No evidence against proportional hazards
-   Small p-value:
    -   Evidence that the PH assumption may be violated
:::

------------------------------------------------------------------------

## Visual Check

### How to read the plot

-   Each panel = one covariate
-   Line shows trend over time

Interpretation:

-   Flat line → PH assumption reasonable\
-   Clear slope → possible violation

```{r}
plot(zph3)
```

------------------------------------------------------------------------

## Key Takeaways

::: fragment
-   Cox model estimates **hazard ratios**
:::

::: fragment
-   No need to specify baseline hazard
:::

::: fragment
-   Interpretation is always **relative risk**
:::

::: fragment
-   Connects directly to KM and log-rank
:::
