---
title: "Coding Lab: Interactions, Nonlinearity, and Model Comparison"
subtitle: "Preparing for Mini-Project #2"
author: "Philip White"
format:
  revealjs:
    slide-number: true
    incremental: true
    theme: serif
    css: slides.css
    scrollable: true
execute:
  echo: true
  warning: false
  message: false
  fig-height: 4
  fig-width: 6
---

```{r setup, include=FALSE}
library(tidyverse)
library(splines)
library(ATbounds)
library(patchwork)

data(RHC)

selected_vars <- c(
  "survival",
  "RHC",
  "age",
  "aps1",
  "surv2md1",
  "pafi1",
  "dnr1_Yes",
  "das2d3pc",
  "chfhx",
  "sex_Female",
  "temp1"
)

RHC_use <- RHC[, selected_vars]

RHC_use <- RHC_use %>%
  mutate(survival = as.integer(as.character(survival)))
```

------------------------------------------------------------------------

## Lab Goals

Today we will discuss:

1.  How we compare models using:
    -   Likelihood ratio tests (`anova`)
    -   AIC / BIC
2.  Discuss why cross-validation is ideal — but why we’ll approximate with AIC/BIC.
3.  Explore interactions in logistic regression.
4.  Explore nonlinear effects using `poly()`, `bs()`, and `ns()`.

This is group-based and exploratory. There is no single correct model.

------------------------------------------------------------------------

## Baseline Model

Start with a simple additive logistic model.

```{r}
fit_base <- glm(survival ~ RHC + age + aps1 + surv2md1 + pafi1 +
                  dnr1_Yes + das2d3pc + chfhx + sex_Female + temp1,
                data = RHC_use,
                family = binomial)

summary(fit_base)
# AIC(fit_base)
# BIC(fit_base)
```

------------------------------------------------------------------------

# Likelihood Ratio Test (LRT)

Used to compare **nested models**:

-   Reduced model: $M_0$
-   Full model: $M_1$ (adds parameters)

Test statistic:

$$
\Lambda = -2 \left[ \ell(M_0) - \ell(M_1) \right]
$$

Under $H_0$:

$$
\Lambda \sim \chi^2_{df}
$$

where $df =$ number of additional parameters.

------------------------------------------------------------------------

## LRT in Logistic Regression

Example:

```{r}
# anova(nested_model, full_model, test = "LRT")
# anova(nested_model, full_model, test = "Chisq") ### equivalent test

```

Interpretation:

-   Small p-value → added terms significantly improve fit.
-   Only valid for **nested models**.
-   Tests statistical improvement, not predictive performance.

------------------------------------------------------------------------

## Information Criteria: AIC

AIC balances - fit (deviance = -2 log-likelihood) - complexity (# of parameters = k):

$$
AIC = -2\ell(\hat\theta) + 2k
$$

where: - $\ell(\hat\theta)$ = maximized log-likelihood - $k$ = number of parameters

Lower AIC = better expected out-of-sample fit.

Example code:

```{r}
AIC(fit_base)
```

------------------------------------------------------------------------

## Why AIC Works

-   Penalizes overfitting.
-   Approximation to leave-one-out cross-validation.
-   Good for predictive modeling.
-   Works for nested or non-nested models.

------------------------------------------------------------------------

## BIC (Bayesian Information Criterion)

Similar to AIC, but stronger penalty:

$$
BIC = -2\ell(\hat\theta) + k \log(n)
$$

Penalty increases with sample size.

Implication:

-   Prefers simpler models.
-   More conservative than AIC.
-   Better for model selection (theory-driven contexts).

Example code:

```{r}
BIC(fit_base)
```

------------------------------------------------------------------------

## Comparing AIC and BIC

| Criterion | Penalty   | Goal                             |
|-----------|-----------|----------------------------------|
| AIC       | $2k$      | Predictive accuracy              |
| BIC       | $k\log n$ | Parsimony / model identification |

As $n$ increases:

-   BIC penalizes complexity more strongly.

------------------------------------------------------------------------

## Cross-Validation: The Gold Standard

Goal: Estimate true out-of-sample performance.

Basic idea:

1.  Split data into training and test sets.
2.  Fit model on training set.
3.  Evaluate performance on test set.
    -   Repeat across folds.

------------------------------------------------------------------------

## K-Fold Cross-Validation

Procedure:

1.  Divide data into $K$ folds.
2.  For each fold:
    -   Train on $K-1$ folds
    -   Test on held-out fold
3.  Average performance across folds.

Common choice: $K = 5$ or $10$.

------------------------------------------------------------------------

## Model Evaluation Metrics (Binary Outcomes)

When evaluating logistic regression models, we can focus on:

1.  **Probability quality**
    -   Log loss
    -   Brier score
2.  **Classification performance**
    -   Sensitivity
    -   Specificity
    -   PPV
    -   NPV
    -   Accuracy
3.  **Ranking ability**
    -   AUC

Different metrics answer different questions.

------------------------------------------------------------------------

## ROC Curve and AUC

::::: columns
::: {.column width="50%"}
**ROC curve plots**

-   **Sensitivity** (true positive rate) on the y-axis\
-   **1 − Specificity** (false positive rate) on the x-axis

**AUC (Area Under the Curve)**

-   Interpretable as the probability a random positive is ranked higher than a random negative.\
-   Threshold-independent measure of **discrimination** (not calibration).\
-   Values: 0.5 = no better than chance; 1.0 = perfect discrimination.
:::

::: {.column width="50%"}
```{r roc_auc_plot, echo=FALSE}
library(pROC)

# compute predicted probabilities
pred_probs <- predict(fit_base, type = "response")
roc_obj <- pROC::roc(response = RHC_use$survival, predictor = pred_probs, quiet = TRUE)

# Build data.frame for ggplot
roc_df <- data.frame(
  fpr = 1 - roc_obj$specificities,
  tpr = roc_obj$sensitivities
)

# AUC and CI
auc_val <- as.numeric(roc_obj$auc)
auc_ci <- pROC::ci.auc(roc_obj)
auc_text <- sprintf("AUC = %.3f\n95%% CI = [%.3f, %.3f]", auc_val, auc_ci[1], auc_ci[3])

# Plot
library(ggplot2)
ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  coord_equal() +
  labs(
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)",
    title = "ROC Curve — Base Logistic Model",
    subtitle = auc_text
  ) +
  theme_minimal(base_size = 13)

```
:::
:::::

------------------------------------------------------------------------

## Summary of Metrics

| Metric      | Requires Threshold? | Measures               |
|-------------|---------------------|------------------------|
| Log Loss    | No                  | Probability accuracy   |
| Sensitivity | Yes                 | True positive rate     |
| Specificity | Yes                 | True negative rate     |
| PPV / NPV   | Yes                 | Predictive value       |
| Accuracy    | Yes                 | Overall classification |
| AUC         | No                  | Ranking ability        |

------------------------------------------------------------------------

## Important Distinction

Different tools answer different questions.

-   **Inference tools** (LRT, p-values) test significance.
-   **AIC/BIC** balance fit and complexity.
-   **Cross-validation metrics** evaluate predictive performance.
-   These tools help us to choose models.

------------------------------------------------------------------------

## Cross-Validation vs AIC/BIC

-   Cross-validation directly estimates predictive performance.

-   AIC approximates expected out-of-sample deviance.

-   BIC approximates model identification under certain assumptions.

-   If computationally feasible: \> Cross-validation is preferred.

But:

-   AIC is fast and often sufficient.
-   BIC is stricter.

------------------------------------------------------------------------

## Practical Guidance

When comparing logistic models:

1.  Use LRT for nested hypothesis testing.
2.  Use AIC/BIC for model selection.
3.  Use cross-validation when prediction is primary goal.
4.  Do not rely on p-values alone.

------------------------------------------------------------------------

## What Is an Interaction?

In a logistic regression model:

$$
\log\left(\frac{p_i}{1-p_i}\right)
= \beta_0 + \beta_1 X_1 + \beta_2 X_2
$$

An **interaction** adds a product term:

$$
\log\left(\frac{p_i}{1-p_i}\right)
= \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2
$$

------------------------------------------------------------------------

## Interpretation With an Interaction

Model:

$$
\log\left(\frac{p_i}{1-p_i}\right)
= \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2
$$

For a one unit change in $X_1$, the effect is:

$$
\frac{\partial}{\partial X_1}
\log\left(\frac{p_i}{1-p_i}\right)
= \beta_1 + \beta_3 X_2
$$

The effect of $X_1$ **depends on the value of** $X_2$.

------------------------------------------------------------------------

## Example: RHC × 2-month Predicted Survival Probability

Model:

$$
\log\left(\frac{p}{1-p}\right)
= \beta_0 + \beta_1 RHC + \beta_2 surv2md1 + \beta_3 RHC \times surv2md1
$$

For patients with RHC = 0: $\log\left(\frac{p}{1-p}\right)= \beta_0 + \beta_2 surv2md1$

For patients with RHC = 1: $\log\left(\frac{p}{1-p}\right) = (\beta_0 + \beta_1) + (\beta_2 + \beta_3) surv2md1$

-   So:
    -   There is a different relationship between log-odds and surv2md1, depending if RHC is 0 or 1.
    -   The treatment effect varies with severity.

------------------------------------------------------------------------

## Interaction Plot on Log-Odds Scale

```{r echo=FALSE}
fit_int <- glm(survival ~ RHC * surv2md1,
               data = RHC_use,
               family = binomial)

surv2md1_seq <- seq(min(RHC_use$surv2md1, na.rm = TRUE),
               max(RHC_use$surv2md1, na.rm = TRUE),
               length.out = 100)

newdat <- expand.grid(
  surv2md1 = surv2md1_seq,
  RHC = c(0,1)
)

newdat$log_odds <- predict(fit_int,
                           newdata = newdat,
                           type = "link")

ggplot(newdat,
       aes(x = surv2md1,
           y = log_odds,
           color = factor(RHC))) +
  geom_line(size = 1.2) +
  labs(
    title = "Interaction on Log-Odds Scale",
    x = "2-Month Survival Probability",
    y = "Log-Odds of Survival",
    color = "RHC"
  ) +
  scale_color_manual(values = c("steelblue","firebrick"),
                     labels = c("No RHC","RHC")) +
  theme_minimal(base_size = 14)
```

------------------------------------------------------------------------

## When Does an Interaction Make Sense?

Because interactions are challenging to interpret, you should generally use interactions when:

-   There is a scientific reason.
-   The effect of one predictor plausibly depends on another.
-   Plots suggest non-parallel trends (interaction plots).
-   Model comparison supports it.

I would be careful using interactions blindly.

------------------------------------------------------------------------

## Questions to ask:

-   Is the interaction clinically meaningful?
-   Does it change predicted probabilities materially?
-   Does it improve interpretability?
-   Does it improve model fit enough to justify complexity?

------------------------------------------------------------------------

## Conceptual Summary

Without interaction:

-   One effect applies to everyone.

With interaction:

-   Effects differ across groups.

In logistic regression:

-   Interaction affects log-odds
-   Interpretation is multiplicative on odds scale

------------------------------------------------------------------------

## Explore Interactions

Pick another interaction that makes substantive sense.

Examples:

-   `age * surv2md1`
-   `sex_Female * das2d3pc`
-   `age * dnr1_Yes`
-   `sex_Female * RHC`
-   Something else?

Fit an interaction model and explain what it means to your group:

```{r}
fit_int <- glm(survival ~ RHC + age + aps1 + surv2md1 + pafi1 +
                  dnr1_Yes + das2d3pc + chfhx + sex_Female + temp1 + age:dnr1_Yes,
               data = RHC_use,
               family = binomial)

summary(fit_int)

anova(fit_base, fit_int, test = "Chisq")
AIC(fit_base, fit_int)
BIC(fit_base, fit_int)
```

------------------------------------------------------------------------

## Explain to group members

-   Is the interaction statistically supported?
-   Can you interpret the interaction parameters?
-   Can you explain the interaction to somebody else?
-   Does AIC or BIC support keeping it?

------------------------------------------------------------------------

## Generalized Additive Model (GAM) Idea

In logistic regression we assume:

$$
\log\left(\frac{p_i}{1-p_i}\right)
=
\beta_0 + \beta_j X_{ij}
$$

That is, the log-odds is **linear** in $X_j$. A GAM allows:

$$
\log\left(\frac{p_i}{1-p_i}\right)=
\beta_0 + f_j(X_{ij}) = 
\beta_0 + \sum_{k=1}^{K} b_k(X_{ij})
$$

where:

-   $b_k(\cdot)$ are basis functions
-   The relationship is smooth but not forced to be linear

------------------------------------------------------------------------

## How We Approximate This in R

We approximate:

$$
f_j(X_j) = \sum_{k=1}^{K} b_k(X_j)
$$

using:

-   `poly(X, d)` → polynomial basis\
-   `bs(X, df)` → B-spline basis\
-   `ns(X, df)` → natural spline basis\
-   Fourier Sum

So we stay within logistic regression, but allow flexible shapes.

------------------------------------------------------------------------

## Nonlinear Effects

Baseline temperature is a variable that works under the "Goldilocks" principle -- the best is not too hot, not too code.

Let’s examine whether nonlinear effects for `temp1` improves the model.

------------------------------------------------------------------------

## Option 1: Polynomial

```{r}
fit_poly <- glm(survival ~ RHC + poly(temp1, 2) + aps1 + age + surv2md1 +
                  pafi1 + dnr1_Yes + das2d3pc + chfhx +
                  sex_Female ,
                data = RHC_use,
                family = binomial)
summary(fit_poly)
```

------------------------------------------------------------------------

## Option 2: Natural Splines

```{r}
fit_ns <- glm(survival ~ RHC +  ns(temp1, df = 3) + aps1 + age + surv2md1 +
                pafi1 + dnr1_Yes + das2d3pc + chfhx +
                sex_Female ,
              data = RHC_use,
              family = binomial)

summary(fit_ns)
```

------------------------------------------------------------------------

## Compare Nonlinear Models

```{r}
anova(fit_base, fit_poly, test = "LRT")
anova(fit_base, fit_ns, test = "LRT")

AIC(fit_base, fit_poly, fit_ns)
BIC(fit_base, fit_poly, fit_ns)
```

------------------------------------------------------------------------

## Explore Nonlinearity

1.  Pick a variable to explore nonlinear relationships.
2.  Fit at least two nonlinear models.
3.  Compare these nonlinear models

------------------------------------------------------------------------

## Final Challenge (Open-Ended)

In your groups:

1.  Propose a “best” model.
2.  Justify:
    -   Interaction choices
    -   Nonlinear choices
    -   Variables included/excluded
3.  Be prepared to defend:
    -   Why not simpler?
    -   Why not more complex?
