---
title: "Ordinal Outcomes: Cumulative Logit Models — NHANES Health Status Lab"
subtitle: "Using `NHANES::NHANES` and `HealthGen` as an ordinal outcome"
author: "Philip White"
format:
  revealjs:
    slide-number: true
    incremental: true
    theme: serif
    css: slides.css
    scrollable: true
    margin: 0
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Minimal packages used in the deck
library(tidyverse)
library(NHANES)    # CRAN package with NHANES sample dataset
library(ordinal)   # clm()
library(broom)
library(patchwork)
library(scales)
theme_set(theme_minimal(base_size = 14))
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Learning goals

-   Define the cumulative logit (proportional odds) model and interpret its coefficients.
-   Understand why `HealthGen` is an ordinal outcome and why that matters.
-   Fit a CLM with `ordinal::clm()` and interpret coefficients (cumulative ORs).
-   Check the proportional odds assumption and consider alternatives.
-   Compare models with LRT / AIC / BIC and visualize predicted probabilities.
-   Optional: brief explanation of NHANES sampling weights and how they'd be used.

------------------------------------------------------------------------

## Logistic Regression (Binary Outcome)

For a binary outcome $Y \in \{0,1\}$, $Y\overset{ind}{\sim}Bern(p_i)$:

$$
\log\!\left(\frac{p_i}{1-p_i}\right)
=
\beta_0 + X_i^\top \beta
$$

------------------------------------------------------------------------

## Logistic Regression via a Latent Variable

-   A useful way to think about logistic regression for binary outcomes: Assume an unobserved continuous variable $S_i = X_i^\top \beta + \varepsilon_i,$ where $\varepsilon_i$ follows a logistic distribution.

-   We don't observe the score but observe: $$
    Y_i =
    \begin{cases}
    1 & \text{if } S_i > 0 \\
    0 & \text{if } S_i \le 0
    \end{cases}
    $$

-   So: $$
    P(Y_i = 1)
    =
    P(S_i > 0)
    =
    P(\varepsilon_i > -X_i^\top \beta) = \frac{e^{X_i^\top \beta}}{1 + e^{X_i^\top \beta}},
    $$ which is the same as logistic regression.

-   Binary logistic regression = a continuous latent variable with a **single cutpoint at 0**.

------------------------------------------------------------------------

## Logistic Regression via a Latent Variable (Visualization)

```{r, fig.width=11, fig.height=6.5, dpi=150, dev='png', out.width='100%', echo=FALSE}

# Create latent variable grid
x <- seq(-6, 6, length.out = 400)

# Logistic density (for illustration)
logistic_density <- dlogis(x)

df_latent <- data.frame(x = x, density = logistic_density)

ggplot(df_latent, aes(x = x, y = density)) +
  geom_line(size = 1.1) +
  geom_area(data = subset(df_latent, x > 0),
            aes(x = x, y = density),
            fill = "steelblue",
            alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", size = 1) +
  labs(
    title = "Latent Variable Interpretation of Logistic Regression",
    x = "Latent score  S = X^Tβ + ε",
    y = "Density"
  ) +
  annotate("text", x = 2.5, y = 0.15,
           label = "Y = 1\n(S > 0)",
           size = 4) +
  annotate("text", x = -2.5, y = 0.15,
           label = "Y = 0\n(S ≤ 0)",
           size = 4) +
  theme_minimal(base_size = 14)
```

------------------------------------------------------------------------

## Cumulative logit model: mathematical statement

For ordered categories **Poor → Fair → Good → Vgood → Excellent**:

$$
\log\!\left(\frac{P(Y \le k)}{P(Y > k)}\right)
=
\theta_k - X^\top \beta
\quad (k = 1,\dots,K-1)
$$

Equivalently,

$$
\log\!\left(\frac{P(Y > k)}{P(Y \le k)}\right)
=
X^\top \beta - \theta_k
$$

### Interpretation

-   $\exp(\beta_j)$ is the multiplicative change in the odds of being in a **higher (better) health category** for a one-unit increase in $X_j$.

-   The effect is the same across all thresholds $k$ (Proportional Odds Assumption).

-   **Key assumption (Proportional Odds).** The same vector $\boldsymbol{\beta}$ applies for every cut $k$. That is, covariate effects shift the latent score equally across all thresholds. Check this assumption in practice (e.g., tests or graphical diagnostics).

------------------------------------------------------------------------

## Latent-Variable Visualization — 5 Ordered Categories (Poor → Excellent)

```{r, fig.width=11, fig.height=6.5, dpi=150, dev='png', out.width='100%', echo=FALSE}

# --- Parameters: adjust these cutpoints to see different mappings ---
cutpoints <- c(-3.77, -1.7, 0.2, 2.02)   # four thresholds -> five categories
cat_labels <- c("Poor", "Fair", "Good", "Vgood", "Excellent")

# Grid for latent variable (logistic latent)
x <- seq(-6, 6, length.out = 2000)
dens <- dlogis(x)   # logistic density

df <- data.frame(x = x, dens = dens)

# Build intervals for shading
breaks <- c(-Inf, cutpoints, Inf)
region_id <- cut(df$x, breaks = breaks, labels = FALSE, right = FALSE)
df$region <- factor(region_id, levels = 1:length(breaks)-1)

# Colors for five categories
fills <- c("#d73027","#fc8d59","#fee08b","#91bfdb","#1a9850") # red -> green ramp

# Compute implied probabilities (area under logistic CDF)
F <- plogis
probs <- numeric(length(breaks)-1)
for (k in 1:(length(breaks)-1)) {
  left <- breaks[k]
  right <- breaks[k+1]
  # Use CDF differences; handle -Inf/Inf
  pl <- if (is.infinite(left) && left < 0) 0 else F(left)
  pr <- if (is.infinite(right) && right > 0) 1 else F(right)
  probs[k] <- pr - pl
}
names(probs) <- cat_labels

# Midpoints for labels (x positions)
mid_x <- sapply(1:(length(breaks)-1), function(k) {
  lx <- max(df$x[df$region == k])
  rx <- min(df$x[df$region == k])
  mean(c(lx, rx))
})
# if any regions empty because of coarse grid, fallback to cutpoint centers:
mid_x[is.na(mid_x)] <- c(-3, (cutpoints[-length(cutpoints)] + cutpoints[-1]) / 2, 3)[is.na(mid_x)]

# Plot
p <- ggplot(df, aes(x = x, y = dens, fill = region)) +
  geom_area(data = subset(df, region == 1), aes(x = x, y = dens), inherit.aes = TRUE) +
  geom_area(data = subset(df, region == 2), aes(x = x, y = dens), inherit.aes = TRUE) +
  geom_area(data = subset(df, region == 3), aes(x = x, y = dens), inherit.aes = TRUE) +
  geom_area(data = subset(df, region == 4), aes(x = x, y = dens), inherit.aes = TRUE) +
  geom_area(data = subset(df, region == 5), aes(x = x, y = dens), inherit.aes = TRUE) +
  scale_fill_manual(values = fills, guide = "none") +
  geom_line(color = "black", size = 0.6) +
  geom_vline(xintercept = cutpoints, linetype = "dashed", size = 0.6) +
  labs(
    title = "Latent Variable Interpretation — Ordered Outcome (5 levels)",
    x = expression(S == X^T * beta + epsilon),
    y = "Density",
    subtitle = paste0("Cutpoints: ", paste0(format(cutpoints, digits=2), collapse = ", "),
                      "  —  Implied probs: ",
                      paste0(cat_labels, ": ", sprintf("%.3f", probs), collapse = " | "))
  ) +
  theme_minimal(base_size = 13)

# add category labels centered in each shaded region
lbl_df <- data.frame(
  x = mid_x,
  y = max(df$dens) * 0.65,
  label = paste0(cat_labels, "\n(", sprintf("%.2f", probs), ")")
)

p + geom_text(
  data = lbl_df,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  size = 3.5,
  color = "black",
  fontface = "bold",
  lineheight = 0.9
)
```

------------------------------------------------------------------------

## Interpretation

-   The latent score $S = X^\top \beta + \varepsilon$ (with $\varepsilon$ following a logistic distribution) is partitioned by thresholds $\theta_1, \dots, \theta_4$ into five observed categories.

-   Category probabilities equal the logistic CDF differences corresponding to the regions between successive thresholds (shown in the subtitle and under each label in the figure).

-   Moving $X$ (or equivalently changing $\beta$) shifts the latent distribution, thereby changing the areas (probabilities) in each region.

-   The cumulative logit model parameterizes the **log-odds of being at or below** each cutpoint: $$\log\left( \frac{P(Y \le k)}{P(Y > k)} \right)=\theta_k - X^\top \beta .$$

-   Equivalently, the log-odds of being **above** the cutpoint are: $$\log\left( \frac{P(Y > k)}{P(Y \le k)} \right)=X^\top \beta - \theta_k .$$

------------------------------------------------------------------------

## About the data: NHANES (package)

We use the `NHANES` package's example dataset (a tidy subset of NHANES suitable for teaching).

-   Outcome: `HealthGen` — self-rated general health (Excellent → Poor).
-   Candidate predictors: BMI, Age, Diabetes status (self-reported), Physical activity indicator, Sex, and Poverty

------------------------------------------------------------------------

## Inspect available columns and preview data

```{r}
# Inspect NHANES dataset included in the package
data("NHANES")   # loads NHANES tibble from the package
glimpse(NHANES, width = 60)

wanted <- c("HealthGen", "BMI", "Age", "Diabetes", "PhysActive","Poverty","SmokeNow","MaritalStatus","Education")

```

------------------------------------------------------------------------

## Create a working analysis dataset

```{r}
NHANES_work <- NHANES %>%
  dplyr::select(all_of(wanted)) %>% 
  mutate(
    HealthGen = factor(HealthGen,
                       levels = c("Poor","Fair","Good","Vgood","Excellent"),
                       ordered = TRUE)
  ) %>% 
    filter(complete.cases(.))

NHANES_work$MaritalStatus = fct_relevel(NHANES_work$MaritalStatus,"Married")
NHANES_work$Education = fct_relevel(NHANES_work$Education,"8th Grade")

# Quick summary
glimpse(NHANES_work)
table(NHANES_work$HealthGen)
```

------------------------------------------------------------------------

## Continuous Exploratory Plot

```{r, fig.width=11, fig.height=6.5, dpi=150, dev='png', out.width='100%', echo=FALSE}
# --- Slide 1: Continuous predictors (boxplots by HealthGen) ---
p_bmi <- ggplot(NHANES_work, aes(x = HealthGen, y = BMI)) +
  geom_boxplot(fill = "cornflowerblue", outlier.size = 0.6) +
  labs(title = "BMI by Self-Reported Health (HealthGen)", x = NULL, y = "BMI") +
  theme(axis.text.x = element_text(angle = 18, hjust = 1))

p_age <- ggplot(NHANES_work, aes(x = HealthGen, y = Age)) +
  geom_boxplot(fill = "lightseagreen", outlier.size = 0.6) +
  labs(title = "Age by Self-Reported Health", x = NULL, y = "Age (years)") +
  theme(axis.text.x = element_text(angle = 18, hjust = 1))

p_pov <- ggplot(NHANES_work, aes(x = HealthGen, y = Poverty)) +
  geom_boxplot(fill = "plum", outlier.size = 0.6) +
  labs(title = "Poverty Index by Self-Reported Health", x = NULL, y = "Poverty Ratio/Score") +
  theme(axis.text.x = element_text(angle = 18, hjust = 1))

slide1_plot <- (p_bmi + p_age + p_pov) + plot_layout(heights = c(1,1))
print(slide1_plot)


```

------------------------------------------------------------------------

## Categorical Exploratory Plot

```{r, fig.width=11, fig.height=6.5, dpi=150, dev='png', out.width='100%', echo=FALSE}
# --- Slide 2: Categorical predictors (proportion within HealthGen) ---

# Diabetes vs HealthGen (proportion)
p_diab <- ggplot(NHANES_work,
                 aes(x = HealthGen,
                     fill = as.factor(Diabetes))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "HealthGen by Diabetes (Proportion)",
       x = NULL,
       y = "Proportion within HealthGen",
       fill = "Diabetes") +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

# Physical Activity vs HealthGen (proportion)
p_phys <- ggplot(NHANES_work,
                 aes(x = HealthGen,
                     fill = as.factor(PhysActive))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "HealthGen by Physical Activity (Proportion)",
       x = NULL,
       y = "Proportion within HealthGen",
       fill = "PhysActive") +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

# Arrange in 1x2 grid
out_cat = (p_diab + p_phys) +
  plot_annotation(title = "Categorical Predictors by Self-Reported Health (Normalized)")

print(out_cat)
```

------------------------------------------------------------------------

## Does ordered structure make sense?

-   `HealthGen` has natural ordering (Poor → Excellent).\

-   Using ordinal model preserves that ordering and gains statistical efficiency vs multinomial model.

-   Group prompt: Why don't we just turn `HealthGen` into a continuous variable?

    -   Are the jumps between Poor→Fair the same as Good→Very Good?

------------------------------------------------------------------------

## Fit a baseline CLM (simple)

```{r}


fit_clm_base <- clm(HealthGen ~ BMI + Age + Diabetes + PhysActive + Poverty,
                    data = NHANES_work,
                    link = "logit",
                    Hess = TRUE)

summary(fit_clm_base)
```

------------------------------------------------------------------------

## Coefficient table and interpretation

```{r}
tidy(fit_clm_base, conf.int = TRUE)
```

-   Each coefficient is a *cumulative* log-odds effect. The change in the log-odds of being in a higher group, holding the other variables constant.

-   $e^{\beta}$ = multiplicative change in odds of being in a **better** health category per unit increase in predictor, holding other variables constant.

-   $100\times(e^{\beta}-1)$ = percent change in odds of being in a **better** health category per unit increase in predictor, holding other variables constant.

-   Example interpretation (fill in with your fitted values):

> -   A one-unit increase in BMI is associated with a x% higher odds of reporting better health (holding other predictors constant).

------------------------------------------------------------------------

## Predicted probabilities: example visualization

::: panel-tabset
### Code

```{r}
# Generate prediction grid
newdat <- expand.grid(
  BMI = seq(18, 40, by = 0.5),
  Age = median(NHANES_work$Age, na.rm = TRUE),
  Diabetes = "Yes",
  PhysActive = "Yes",
  Poverty = median(NHANES_work$Poverty, na.rm = TRUE)
)

# Compute predicted probabilities
preds <- predict(fit_clm_base, newdata = newdat, type = "prob")$fit

pred_df <- cbind(newdat, preds) %>%
  tidyr::pivot_longer(
    -names(newdat),
    names_to = "category",
    values_to = "prob"
  )

head(pred_df)
```

### Plot

```{r, echo=FALSE, fig.width=11, fig.height=6.5, dpi=150, dev='png', out.width='100%'}
ggplot(pred_df, aes(x = BMI, y = prob, color = category)) +
  geom_line(size = 1.2) +
  labs(
    title = "Predicted HealthGen Probabilities vs BMI",
    y = "Predicted probability",
    x = "BMI"
  ) +
  theme_minimal(base_size = 16)
```
:::

------------------------------------------------------------------------

## Competing Model

```{r}
fit_clm_complex <- clm(HealthGen ~ BMI + Age + Diabetes + PhysActive + 
                      Poverty + SmokeNow + MaritalStatus + Education,
                    data = NHANES_work,
                    link = "logit",
                    Hess = TRUE)
summary(fit_clm_complex)
```

------------------------------------------------------------------------

## Model comparison: nested models and LRT

Compare a the more complex model to the base model:

```{r}
anova(fit_clm_base,fit_clm_complex)
```

-   LRT tests whether `SmokeNow + MaritalStatus + Education` adds explanatory power (assuming nested models).

------------------------------------------------------------------------

## AIC / BIC for model selection

```{r}
AIC(fit_clm_base,fit_clm_complex)
BIC(fit_clm_base,fit_clm_complex)
```

-   Lower AIC/BIC indicates preferable model (trade-off between fit and complexity).
-   AIC usually preferred for predictive focus; BIC penalizes complexity more.

------------------------------------------------------------------------

## Cross-validation (conceptual notes)

-   K-fold CV or repeated CV gives an empirical estimate of predictive performance.
-   For ordinal outcomes, deviance or log-loss (negative log-likelihood) can be used as scoring metric; probability-based scoring is preferred over accuracy.

Example (conceptual code sketch — not executed here):

```{r}
# Pseudocode: (do not run as-is)
# for (fold in folds) {
#   train <- data[fold != k, ]
#   test  <- data[fold == k, ]
#   fit <- clm(..., data = train)
#   preds <- predict(fit, newdata = test, type = "prob")
#   compute log-loss or deviance on test
# }
```

------------------------------------------------------------------------

## Proportional odds assumption

-   The model assumes the effect of a predictor is the same across all cumulative logits.
-   If violated, a single $\beta$ is not appropriate to control movement across ordinal bins.

Check with a nominal test:

```{r}
nominal_test(fit_clm_complex)
```

Interpretation:

-   Large p-value: PO assumption not rejected → CLM plausible.
-   Small p-value: PO assumption may be violated; consider alternatives.

------------------------------------------------------------------------

## If PO assumption fails — options

1.  Partial proportional odds models (Add nominal effects for some variables)
2.  Adjacent-category or continuation-ratio models\
3.  Multinomial logistic regression (no ordering used)

Discuss trade-offs: interpretability vs flexibility.

------------------------------------------------------------------------

## Lab: group exercise (open-ended)

In groups (30–45 minutes):

1.  Fit a baseline CLM (like above).
2.  Explore nonlinearity for BMI (e.g., `poly(BMI,2)`, `ns(BMI, df=4)`), compare with AIC/LRT.
3.  Test interactions you think make sense (e.g., `Sex_f:BMI`, `Age:PhysAct_f`) and interpret.
4.  Check proportional odds assumption.
5.  Choose a final model and justify (interpret 2 coefficients in plain language).
6.  Plot predicted probabilities for two contrasting profiles and discuss.

------------------------------------------------------------------------

## Discussion prompts

-   Does accounting for ordering change conclusions vs treating health as categorical?
-   How would sampling weights change the questions you answer?
-   Which predictors appear most clinically relevant?
-   When would you prefer a multinomial model despite losing efficiency?

------------------------------------------------------------------------

## Takeaways

-   `HealthGen` is a classic ordinal outcome — CLM is a natural model.
-   Interpret coefficients as cumulative odds ratios.
-   Always check the proportional odds assumption.
-   Model selection requires balancing predictive performance and interpretability.
-   Visualizing predicted probabilities is essential for communicating results.
