---
title: "Survival Analysis — Probability, Life Tables, and KM-Curves"
author: "Philip White"
format:
  revealjs:
    slide-number: true
    incremental: true
    theme: serif
    css: slides.css
    scrollable: true
    margin: 0
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
```

------------------------------------------------------------------------

## Lecture 1 — Roadmap

```         
1.  Basic probability view: CDF, survival function, hazard function.
2.  Life tables (discrete / interval estimates) — conditional probs.
3.  Right censoring: what breaks and what doesn’t.
4.  Kaplan–Meier estimator: derivation, hand calculations, and R code.
5.  Quick diagnostics and interpretation.
```

------------------------------------------------------------------------

## Motivation: Why Survival Analysis?

Many biomedical questions involve **time until an event**.

Examples:

-   Time from diagnosis to death

-   Time until relapse

-   Time until infection

-   Time until device failure

-   Key feature:

> -   The outcome is **time until the event**, not simply whether the event occurs.

------------------------------------------------------------------------

## What makes time-to-event data different?

Two important characteristics:

1.  **Events occur over time**
2.  **Some subjects may not experience the event during follow-up**

Example situations:

-   Patient still alive at last follow-up
-   Study ends before event occurs
-   Patient lost to follow-up

These features require specialized statistical tools.

------------------------------------------------------------------------

## Probability View of Survival Data

Let $T$ be the **time to event**.

We can treat $T$ as a random variable.

We'll introduce four key functions describe the distribution of $T$.

------------------------------------------------------------------------

## Cumulative Distribution Function

The cumulative distribution function (CDF) is

$$
F(t) = P(T \le t)
$$

Interpretation:

> The probability that the event occurs **on or before time** $t$.

------------------------------------------------------------------------

## Survival Function

The survival function is

$$
S(t) = P(T > t)
$$

This is the probability that the event **has not yet occurred by time** $t$.

Relationship with the CDF:

$$
S(t) = 1 - F(t)
$$

This function is the **central quantity in survival analysis**.

------------------------------------------------------------------------

## Probability Density

If $T$ has a density:

$$
f(t) = \frac{d}{dt}F(t)
$$

The density describes how event times are distributed across time.

Relationship with survival:

$$
f(t) = -\frac{d}{dt}S(t)
$$

------------------------------------------------------------------------

## Hazard Function

The hazard function measures the **instantaneous risk of the event**.

Definition:

$$
h(t) =
\lim_{\Delta t \to 0}
\frac{P(t < T \le t + \Delta t \mid T \ge t)}{\Delta t}
$$

Equivalent expression:

$$
h(t) = \frac{f(t)}{S(t)}
$$

Interpretation:

> The instantaneous rate at which events occur at time $t$, given survival up to $t$.

------------------------------------------------------------------------

## Intuition for the Hazard

Think of the hazard as:

-   the **instantaneous event rate**
-   conditional on having survived to time $t$

Important:

-   It is **not a probability**
-   It can exceed 1
-   Units are events per unit time

------------------------------------------------------------------------

## Relationships Among Functions

These functions are mathematically connected:

$$
S(t) = 1 - F(t)
$$

$$
f(t) = \frac{d}{dt}F(t)
$$

$$
h(t) = \frac{f(t)}{S(t)}
$$

Also,

$$
S(t) =
\exp\left(-\int_0^t h(u) \, du\right)
$$

This link between hazard and survival will become important later.

------------------------------------------------------------------------

## Example: Exponential Survival - Find the CDF

Suppose

$$
f(t) = \lambda e^{- \lambda t}
$$

Find the CDF $F(t) = \int_{-\infty}^t f(t) dt$.

------------------------------------------------------------------------

## Example: Exponential Survival - find the hazard function

Then the survival function is

$$
S(t) = e^{-\lambda t}
$$

-   Find the hazard function $h(t) = \frac{f(t)}{S(t)}$. Explain what this means to your group.

Properties:

-   constant hazard
-   decreasing survival
-   mean survival time $1/\lambda$\$

------------------------------------------------------------------------

## Visualizing Exponential Survival

```{r}
lambda <- 0.1
t <- seq(0,50,length.out=400)

S <- exp(-lambda * t)

ggplot(tibble(t,S), aes(x=t,y=S)) +
  geom_line() +
  labs(
    x="Time",
    y="Survival probability",
    title="Exponential survival curve"
  )
```

------------------------------------------------------------------------

## Summary

Key ideas:

-   Survival analysis studies **time until an event**
-   The key functions are:
    -   $F(t)$: cumulative distribution
    -   $S(t)$: survival function
    -   $f(t)$: density
    -   $h(t)$: hazard
-   Survival curves are often **step functions** when estimated empirically

------------------------------------------------------------------------

## Empirical Survival Curves

Before introducing formal models, we can look at **empirical survival curves**.

Suppose we observe the following event times:

| Person | Event Time |
|--------|------------|
| A      | 2          |
| B      | 4          |
| C      | 6          |
| D      | 8          |
| E      | 10         |

At time $t$, survival probability can be estimated as:

$$
\hat S(t) =
\frac{\text{number of individuals with event time } > t}{n}
$$

------------------------------------------------------------------------

## Example: Empirical Survival Curve

::: panel-tabset
### Data

| Person | Event Time |
|--------|------------|
| A      | 2          |
| B      | 4          |
| C      | 6          |
| D      | 8          |
| E      | 10         |

Can you draw the empirical survival function?

$$
\hat S(t) =
\frac{\text{number of individuals with event time } > t}{n}
$$

### Survival Function

For the previous data:

| Time | Alive | Survival |
|------|-------|----------|
| 0    | 5     | 1.0      |
| 2    | 4     | 0.8      |
| 4    | 3     | 0.6      |
| 6    | 2     | 0.4      |
| 8    | 1     | 0.2      |
| 10   | 0     | 0        |

Survival curves are **step functions**.

### Plot

```{r}
times <- c(2,4,6,8,10)

df <- tibble(time = times)

ggplot(df, aes(x = time)) +
  stat_ecdf(geom = "step", aes(y = 1 - after_stat(y))) +
  labs(
    x = "Time",
    y = "Empirical survival estimate",
    title = "Empirical survival curve"
  ) + 
  scale_x_continuous(limits = c(0,12),breaks = seq(0,12,by = 2))
```

This shows the **empirical survival function** based purely on observed event times.
:::

------------------------------------------------------------------------

## Variance of the Empirical Survival Estimate

Suppose we estimate survival empirically:

$$
\hat S(t) = \frac{\text{# individuals with } T > t}{n}
$$

Let

-   $n$ = total number of individuals
-   $k$ = number of events observed by time $t$

Then

$$
\hat S(t) = \frac{n-k}{n}
$$

Because this is a **sample proportion**, the variance is

$$
\mathrm{Var}(\hat S(t)) =
\frac{\hat S(t)\left(1-\hat S(t)\right)}{n}
$$

Standard error:

$$
SE(\hat S(t)) =
\sqrt{\frac{\hat S(t)(1-\hat S(t))}{n}}
$$

Approximate 95% confidence interval:

$$
\hat S(t) \pm 1.96 \times SE(\hat S(t))
$$

------------------------------------------------------------------------

## Life Tables

::: fragment
Suppose we follow individuals over time and record:
:::

::: fragment
-   Number **at risk** at the start of each interval
-   Number of **events (deaths)** during the interval
:::

::: fragment
Life tables estimate survival **interval-by-interval**.
:::

::: fragment
Example structure:
:::

::: fragment
| Year of Treatment | Patients Treated | 1st Year | 2nd Year | 3rd Year | 4th Year | 5th Year |
|---:|---:|---:|---:|---:|---:|---:|
| 2000 | 84 | 44 | 21 | 13 | 10 | 8 |
| 2001 | 62 | 31 | 14 | 10 | 6 |  |
| 2002 | 93 | 50 | 20 | 13 |  |  |
| 2003 | 60 | 29 | 16 |  |  |  |
| 2004 | 76 | 43 |  |  |  |  |
| **Column Total** | **375** | **197** | **71** | **36** | **16** | **8** |
:::

::: fragment
For each interval we compute the **conditional survival probability**.
:::

## Conditional Probability View of Survival

::: fragment
Let

-   $n_j$ = number at risk at start of interval $j$
-   $d_j$ = number of events during interval $j$
:::

::: fragment
The probability of surviving that interval is

$$
p_j = 1 - \frac{d_j}{n_j}
$$
:::

::: fragment
This is the **conditional probability of surviving the interval given survival up to its start**.
:::

------------------------------------------------------------------------

## Building the Survival Curve

::: fragment
We combine conditional probabilities.

If

$$
p_1, p_2, p_3, \dots
$$

are the interval survival probabilities, then

$$
S(t_j) = p_1 p_2 \cdots p_j
$$
:::

::: fragment
So survival is the **product of conditional survival probabilities**.

This idea is the foundation for **life tables and Kaplan–Meier curves**.
:::

------------------------------------------------------------------------

## Example: Life Table Calculation

::: fragment
Suppose we observe:

| Year of Treatment | Patients Treated | 1st Year | 2nd Year | 3rd Year | 4th Year | 5th Year |
|---:|---:|---:|---:|---:|---:|---:|
| 2000 | 84 | 44 | 21 | 13 | 10 | 8 |
| 2001 | 62 | 31 | 14 | 10 | 6 |  |
| 2002 | 93 | 50 | 20 | 13 |  |  |
| 2003 | 60 | 29 | 16 |  |  |  |
| 2004 | 76 | 43 |  |  |  |  |
| **Column Total** | **375** | **197** | **71** | **36** | **16** | **8** |
:::

::: fragment
Compute conditional survival:

| Interval | At Risk  | Alive at End | $p_j$                   |
|----------|----------|--------------|-------------------------|
| Year 1   | 375      | 197          | $197/375 = 0.525$       |
| Year 2   | 197 - 43 | 71           | $71/(197 - 43) = 0.461$ |
| Year 3   | 71 - 16  | 36           | $36/(71 - 16) = 0.655$  |
| Year 4   | 36 - 13  | 16           | $16/(36 - 13) = 0.696$  |
| Year 5   | 16 - 6   | 8            | $8/(16 - 6) = 0.800$    |
:::

------------------------------------------------------------------------

## Life Table Survival Curve

Survival estimates:

| Time   | Survival                     |
|--------|------------------------------|
| 0      | 1.00                         |
| Year 1 | $0.525$                      |
| Year 2 | $0.525 \times 0.461 = 0.242$ |
| Year 3 | $0.242 \times 0.655 = 0.158$ |
| Year 4 | $0.158 \times 0.696 = 0.110$ |
| Year 5 | $0.110 \times 0.800 = 0.088$ |

Interpretation:

::: fragment
> Approximately **16% survive past year 3**.
:::

::: fragment
> Approximately **9% survive past year 5**.
:::

::: fragment
Life tables estimate survival using **interval-based conditional probabilities**.
:::

------------------------------------------------------------------------

## The Problem with Real Data

Real studies often have **incomplete follow-up**.

Example:

| Person | Time | Event                  |
|--------|------|------------------------|
| A      | 2    | death                  |
| B      | 4    | death                  |
| C      | 6    | withdrew study consent |
| D      | 7    | death                  |
| E      | 9    | moved to Australia     |

-   We know the person survived **at least until that time**
-   But we do not know what happened afterward.'=
-   This data is **censored**?

------------------------------------------------------------------------

## Right Censoring

::: fragment
Right censoring occurs when

$$
T > C
$$

where $C$ is the last time we observe the individual.
:::

::: fragment
Examples:

-   Study ends
-   Patient lost to follow-up
-   Participant withdraws
:::

::: fragment
Censored observations still contribute information **before the censoring time**.
:::

------------------------------------------------------------------------

## Assumptions About Censoring

Kaplan–Meier estimation works under several key assumptions.

::: fragment
### 1. Independent Censoring

Censoring must be **independent of the event time**.

Formally,

$$
P(\text{censored at } t \mid T \ge t)
\text{ does not depend on the future event time.}
$$
:::

::: fragment
Interpretation:

> Individuals who are censored at time $t$ have the same future risk as those who remain under observation.
:::

------------------------------------------------------------------------

### 2. Non-informative Censoring

::: fragment
The reason for censoring should **not be related to the outcome**.
:::

::: fragment
Examples that are usually acceptable:

-   Study ends before the event occurs
-   Administrative censoring at a fixed date
-   Patient moves away for unrelated reasons
:::

::: fragment
Examples that violate the assumption:

-   Patients drop out because their condition worsens
-   Patients withdraw due to severe side effects
:::

------------------------------------------------------------------------

### 3. Exact Event Times Are Known

::: fragment
Kaplan–Meier assumes:

-   Event times are measured accurately
-   Ties may occur but the ordering of events is known
:::

::: fragment
### Why This Matters

If censoring is informative:

-   Kaplan–Meier curves may be **biased**
-   Survival may be **overestimated or underestimated**
:::

::: fragment
This assumption cannot be tested directly from the data and must be justified by study design.
:::

------------------------------------------------------------------------

## Example of Assumption Violation

::: fragment
Suppose very sick patients are more likely to drop out of the study. What direction would the Kaplan–Meier survival estimate be biased?
:::

::: fragment
Is survival underestimate or overestimated?
:::

------------------------------------------------------------------------

## Other Problem with Life Tables

Life tables use **time intervals**, but we often observe **exact event times**.

::: fragment
Example:

| Person | Event Time |
|--------|------------|
| A      | 2          |
| B      | 4          |
| C      | 7          |
| D      | 9          |
:::

::: fragment
Grouping these into intervals loses information. So, we want a method that uses **exact event times**.
:::

------------------------------------------------------------------------

## Kaplan–Meier (KM) Curve

Kaplan–Meier estimates survival using **each event time directly**.

Let

-   $t_1 < t_2 < \dots < t_k$ be event times (ties are okay)
-   $n_j$ = number at risk just before $t_j$
-   $d_j$ = number of events at $t_j$

::: fragment
Then

$$
\widehat{S(t)} =
\prod_{t_j \le t}
\left(1 - \frac{d_j}{n_j}\right)
$$

-   This is called the **product-limit estimator**.
:::

------------------------------------------------------------------------

## Key Idea of Kaplan–Meier

::: fragment
Kaplan–Meier is simply the **same conditional probability idea**:

1.  Compute survival probability at each event time
2.  Multiply the probabilities together
:::

::: fragment
The only difference from life tables:

> We use **exact event times instead of intervals**.
:::

------------------------------------------------------------------------

## Simple Kaplan–Meier Example

Suppose we observe:

| Time | Event    |
|------|----------|
| 2    | death    |
| 4    | death    |
| 6    | censored |
| 7    | death    |
| 10   | censored |

::: fragment
Step 1: identify event times.

Step 2: number at risk at times of events.
:::

::: fragment
| Time | At Risk |
|------|---------|
| 2    | 5       |
| 4    | 4       |
| 7    | 2       |
:::

------------------------------------------------------------------------

## KM Calculation

::: fragment
Step 3. Compute conditional survival at 2, 4, and 7:
:::

::: fragment
| Time | Survival Step    |
|------|------------------|
| 2    | $1 - 1/5 = 0.8$  |
| 4    | $1 - 1/4 = 0.75$ |
| 7    | $1 - 1/2 = 0.5$  |
:::

------------------------------------------------------------------------

## Kaplan–Meier Survival

Step 4. Now build KM-curve. Multiply conditional survival.

::: fragment
| Time | Survival                 |
|------|--------------------------|
| 0    | 1.00                     |
| 2    | 0.8                      |
| 4    | $0.8 \times 0.75 = 0.60$ |
| 7    | $0.60 \times 0.50 = 0.3$ |

Interpretation:

> About **30% survive beyond time 7**.
:::

------------------------------------------------------------------------

## Example in R

::::: panel-tabset
### Code

::: fragment
-   `event = 1` → death occurred
-   `event = 0` → right censored
:::

::: fragment
```{r,eval=FALSE}
library(survival)
library(survminer)


time <- c(2,4,6,7,10)
event <- c(1,1,0,1,0)
data_use = Surv(time, event)

fit <- survfit(Surv(time, event) ~ 1)

out = suppressWarnings(ggsurvplot(fit, 
           data = data_use,
           risk.table = TRUE,
           conf.int = FALSE))

out$plot

```
:::

### Plot

```{r,echo = FALSE}
library(survival)
library(survminer)


time <- c(2,4,6,7,10)
event <- c(1,1,0,1,0)
data_use = Surv(time, event)

fit <- survfit(Surv(time, event) ~ 1)

out = suppressWarnings(ggsurvplot(fit, 
           data = data_use,
           risk.table = TRUE,
           conf.int = FALSE))

out$plot

```
:::::

------------------------------------------------------------------------

## Practice Group Problem — Compute a KM Curve by Hand

::: fragment
| Person | Time | Event    |
|--------|------|----------|
| A      | 1    | death    |
| B      | 3    | death    |
| C      | 4    | censored |
| D      | 5    | death    |
| E      | 6    | death    |
| F      | 8    | censored |
:::

::: fragment
1.  Identify the **event times**.
2.  Determine **how many are at risk** at each event time.
3.  Compute the **conditional survival probabilities**.
4.  Build the KM curve step-by-step.
:::

------------------------------------------------------------------------

## Steps 1 and 2 — Risk Set

::: fragment
Fill in the number **at risk just before each event time**.

| Time | Event? | At Risk |
|------|--------|---------|
| 1    | death  | ?       |
| 3    | death  | ?       |
| 5    | death  | ?       |
| 6    | death  | ?       |
:::

::: fragment
Remember:

-   Remove individuals **after they die**
-   Remove individuals **after censoring**
-   Censored individuals still count **before the censoring time**
:::

::: fragment
Why does the censored observation at time 4 affect the risk set at time 5?
:::

------------------------------------------------------------------------

## Step 3 — KM Conditional Survival

Now compute

$$
1 - \frac{d_j}{n_j}
$$

| Time | At Risk | Deaths | Survival Step |
|------|---------|--------|---------------|
| 1    | ?       | 1      | ?             |
| 3    | ?       | 1      | ?             |
| 5    | ?       | 1      | ?             |
| 6    | ?       | 1      | ?             |

------------------------------------------------------------------------

## Step 4: KM Curve

::: fragment
Then compute cumulative survival:

$$
S(t_j) = \prod_{k \le j} \left(1 - \frac{d_k}{n_k}\right)
$$
:::

::: fragment
1.  Compute the cumulative survival values
2.  Draw the Kaplan–Meier curve

-   When does the curve drop?
-   When does it stay flat?
-   Where would the **censoring marks** appear?
:::

------------------------------------------------------------------------

## Shape of a Kaplan–Meier Curve

Kaplan–Meier curves are **step functions**.

-   They drop **only at event times**
-   They remain flat between events
-   Censored observations appear as **tick marks**

------------------------------------------------------------------------

## Summary

Key ideas:

-   Survival curves can be built from **conditional probabilities**
-   Life tables estimate survival using **intervals**
-   Kaplan–Meier uses **exact event times**
-   Censored observations contribute information **up to the censoring time**

Next lecture:

-   Comparing survival curves
-   **Log-rank tests**
-   Stratified survival analysis
